\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{κ}{\ensuremath{\kappa}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{ε}{\ensuremath{\varepsilon}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\$\kappa_3$ asymptotic for the anharmonic Gibbs}
\author{Daniel Murfet}
\date{6 May 2026}

\begin{document}
\maketitle

\begin{abstract}
This retrospective records the first \emph{cross-repo} Tide-loop
excursion: a single theorem stitching together the abstract $\kappa_3$
framework from \texttt{timaeus-research/threepoint} and the concrete
anharmonic asymptotic machinery from \texttt{timaeus-research/laplace},
and proving that for the 1D anharmonic potential
$L(x) = (\lambda/2)x^2 + (\alpha/6)x^3 + (\gamma/24)x^4$ with
$\lambda, \gamma > 0$ and the discriminant condition $\alpha^2 < 3\lambda\gamma$,
the third cumulant of the linear observable in all three slots admits
the leading asymptotic
$t^2 \kappa_3(x, x, x) \to -\alpha/\lambda^3$. The formalisation took
325 lines of Lean in a new file \texttt{Laplace/OneD/AnharmonicKappa3.lean},
with zero \texttt{sorry} / \texttt{axiom} / \texttt{native\_decide}, and
landed in a single session. The decisive contributions of the
deliberation step were a numerical sanity check that caught a
miscomputed Wick coefficient (Claude wrote $\langle u^6 \rangle_0 \cdot \alpha/6 = 3\alpha/2$
when it is in fact $5\alpha/2$), and an architectural recommendation
from GPT-5.5~Pro to add \texttt{threepoint} as a Lake dependency of
\texttt{laplace}, keeping the dependency direction
generic-to-analytic. The headline result is the actual cross-susceptibility
identity the threepoint paper was written for: the cumulant
$\kappa_3$ at leading order is non-zero precisely because the cubic
perturbation breaks the Gaussian parity that makes the harmonic case
trivially zero.
\end{abstract}

\section{Setting}

The threepoint repository, after Tide~5 (kappa3-harmonic), instantiated
its abstract $\kappa_3$ framework on the harmonic Gibbs measure and
proved the trivial vanishing
$\kappa_3(x, x, x) = 0$ by parity. That tide deliberately avoided the
\texttt{GibbsRegularity} hypothesis bundle\footnote{It bundles partition
positivity, differentiability of the unperturbed partition at $h = 0$,
and the integral-of-derivative identities. Required for the FDT and
cross-susceptibility \emph{derivative} theorems but not for value-at-zero
identities.} because $\kappa_3$ is defined at $h = 0$, and at the
harmonic level every term in the cumulant expansion contains an odd-power
Gaussian moment.

The Tide~5 retrospective flagged the natural next step as
``$\kappa_3$ for the anharmonic'': the cubic term $(\alpha/6)x^3$ in
the anharmonic potential breaks parity, so $\kappa_3$ is genuinely
non-zero, and the formalisation has to do an actual asymptotic
calculation rather than read off Wick contractions. The candidates
survey of 6~May 2026 listed this as \textbf{C2/F1}, a
``substantively cross-repo bridge'' --- the formalisation lives at the
boundary between the two seabeds.

The laplace repository, by 6~May, had a long history of asymptotic
expansions for the anharmonic Gibbs measure, headlined by
\texttt{cov\_anharmonic\_asymptotic}\footnote{From
\texttt{Laplace/OneD/IntegralRemainder.lean}; the original 27--29~April
formalisation.}: the first-order asymptotic
$\mathrm{Cov}_t[x^2, x] = -2\alpha/(\lambda^3 t^2) + o(t^{-2})$
for the same anharmonic potential under the same discriminant condition.
Underneath sat the rescaling identity, the J-function asymptotics, and a
constellation of polynomial-times-Gaussian moment lemmas. The candidates
survey noted that the moment expansions $\langle x \rangle_t,
\langle x^2 \rangle_t, \langle x^3 \rangle_t$ ought to have leading-order
forms that combine cleanly with $\mathrm{Cov}_t[x^2, x]$ to give the
$\kappa_3$ asymptotic.

\section{The thing formalised}

The headline statement is:

\begin{quote}\itshape
For $\lambda > 0$, $\gamma > 0$, $\alpha^2 < 3\lambda\gamma$, the
1D anharmonic potential $L(x) = (\lambda/2)x^2 + (\alpha/6)x^3 + (\gamma/24)x^4$
on $\mathbb{R}$ against Lebesgue measure, and the linear observable
$\varphi(x) = A(x) = B(x) = x$ in all three slots,
\[
  t^2 \cdot \kappa_3\!\big(\mathrm{vol}, L, x, t, x, x\big)
  \;\longrightarrow\; -\frac{\alpha}{\lambda^3}
  \qquad \text{as } t \to \infty.
\]
\end{quote}

The Lean signature (in
\texttt{Laplace/OneD/AnharmonicKappa3.lean}\footnote{Permalink:
\texttt{Laplace/OneD/AnharmonicKappa3.lean} on
\texttt{tide/anharmonic-kappa3} at commit \texttt{8a62893}.}):

\begin{verbatim}
theorem kappa3_anharmonic_id_id_id_asymptotic
    {lam alpha gamma : ℝ}
    (hlam : 0 < lam) (hgamma : 0 < gamma)
    (hdisc : alpha ^ 2 < 3 * lam * gamma) :
    Filter.Tendsto
      (fun t : ℝ => t ^ 2 * Threepoint.kappa3 (volume : Measure ℝ)
          (anharmonicPotential lam alpha gamma)
          (fun x : ℝ => x) t (fun x : ℝ => x) (fun x : ℝ => x))
      Filter.atTop
      (nhds (-alpha / lam ^ 3))
\end{verbatim}

The discriminant condition $\alpha^2 < 3\lambda\gamma$ is the same one
that governs the existing covariance
asymptotic.\footnote{\texttt{cov\_anharmonic\_asymptotic} in
\texttt{IntegralRemainder.lean}.} It ensures $L$ is globally coercive,
so the Gibbs measure is well-defined and the Laplace asymptotic is
dominated by $x = 0$. The hypothesis was the load-bearing correction
from a much earlier GPT consult on the original anharmonic
formalisation; here it is inherited unchanged.

The instantiation $\varphi = A = B = x$ in all three slots of $\kappa_3$
matches the harmonic statement from
Tide~5:\footnote{\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero}.}
the ``id, id, id'' naming refers to all three observables, including
the perturbation observable $A$, being the identity. (A subtle
conceptual point caught mid-formalisation: the $A$ parameter does not
drop out of the cumulant expansion at $h = 0$, because $A$ also appears
as a normal observable inside terms like $\langle A \cdot B \rangle \cdot \langle \varphi \rangle$.
See \S\ref{sec:roadblocks}.)

\section{Proof strategy}\label{sec:strategy}

The proof factors through a polynomial identity in the moments and
three asymptotic theorems already proven in laplace. The chain of
reasoning is short enough to write out:

\paragraph{The algebraic identity.} For any measure and any potential,
\[
  \kappa_3(x, x, x)
  \;=\; \mathrm{Cov}[x^2, x] - 2 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3.
\]
Verification by expansion: $\mathrm{Cov}[x^2, x] = \langle x^3 \rangle - \langle x^2 \rangle \langle x \rangle$,
so the right-hand side is
$\langle x^3 \rangle - \langle x^2 \rangle \langle x \rangle - 2 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3
= \langle x^3 \rangle - 3 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3 = \kappa_3(x, x, x)$.
This is a pure ring identity in the three moments; in Lean it is one
\texttt{ring} call after expanding \texttt{gibbsCov}.

\paragraph{The three asymptotics.} All three exist as user-facing
theorems in laplace's integral-remainder module.\footnote{Path:
\texttt{Laplace/OneD/IntegralRemainder.lean}. See
\texttt{retrospective.tex} (\textsection 6--7) for the originals.}
\begin{itemize}
\item \texttt{cov\_anharmonic\_asymptotic}: $t^2 \cdot \mathrm{Cov}_t[x^2, x] \to -2\alpha/\lambda^3$.
\item \texttt{mean\_anharmonic\_asymptotic}: $t \cdot \langle x \rangle_t \to -\alpha/(2\lambda^2)$.
\item \texttt{cov\_self\_anharmonic\_asymptotic}: $t \cdot \mathrm{Var}[x]_t \to 1/\lambda$
(stated as $\mathrm{Cov}_t[x, x]$, i.e.\ the variance).
\end{itemize}

\paragraph{One small new asymptotic.} The bridge needs $t \cdot \langle x^2 \rangle_t \to 1/\lambda$,
which is not directly stated in laplace but follows from the existing
variance and mean asymptotics via $\langle x^2 \rangle = \mathrm{Var}[x] + \langle x \rangle^2$.
The mean-square term $\langle x \rangle^2$ is $O(1/t^2)$ (since $t \cdot \langle x \rangle$
is bounded), so $t \cdot \langle x \rangle^2$ is $O(1/t) \to 0$; combine
with $t \cdot \mathrm{Var}[x] \to 1/\lambda$ via \texttt{Tendsto.add}.
This is packaged as \texttt{secondMoment\_anharmonic\_asymptotic}, ~25
lines, exposed at the top level for downstream reuse.

\paragraph{Combining.} The headline theorem unfolds the seven
\texttt{Threepoint.gibbsExp~...~0} terms inside $\kappa_3$ to
\texttt{Laplace.gibbsExpectation} (one bridging lemma; see
\S\ref{sec:roadblocks}), applies the algebraic identity, and combines
the three asymptotics via \texttt{Tendsto.add}, \texttt{Tendsto.mul},
and \texttt{Tendsto.const\_mul}. The terminal limit-value computation
is one \texttt{field\_simp + ring}:
\[
  -\tfrac{2\alpha}{\lambda^3} - 2 \cdot \tfrac{1}{\lambda} \cdot \big(-\tfrac{\alpha}{2\lambda^2}\big) + 0
  \;=\; -\tfrac{2\alpha}{\lambda^3} + \tfrac{\alpha}{\lambda^3}
  \;=\; -\tfrac{\alpha}{\lambda^3}.
\]
The mathematical content is roughly 30~lines; the rest of the 325-line
file is the four bridging lemmas and the \texttt{Tendsto}-algebra
plumbing.

\section{Roadblocks and resolutions}\label{sec:roadblocks}

\paragraph{The miscomputed Wick coefficient.} Pre-deliberation Claude
calculated
\[
  \langle x^3 \rangle_t \approx -\tfrac{3\alpha}{2 \lambda^3 t^2},
\]
which gives a $\kappa_3$ leading order of $0 \cdot t^{-2} + O(t^{-3})$
--- a vanishing claim. The actual coefficient is
$5\alpha/(2\lambda^3 t^2)$, because $\langle u^6 \rangle_0 = 5!! = 15$
in the rescaled Gaussian, giving the prefactor $15/6 = 5/2$ rather than
$3/2$. The leading-order $\kappa_3$ is then $-\alpha/\lambda^3 \cdot t^{-2}$,
\emph{not} zero.

The error was caught \emph{before} the GPT consult by running a small
\texttt{scipy.integrate.quad} sanity check\footnote{Saved at
\texttt{sri/projects/patterning/staging/anharmonic-kappa3-numerics.py}.}
at $\lambda = 1, \alpha = 0.5, \gamma = 1$ and $t \in \{4, 8, \dots, 256\}$.
The numerical $\kappa_3 \cdot t^2$ converges to $-0.495$, matching the
corrected prediction $-\alpha/\lambda^3 = -0.5$ to four significant
figures. GPT-5.5~Pro independently arrived at the same correction with
the same $5/2$ vs.\ $3/2$ diagnosis, which gave a high-confidence
double-check before any Lean was attempted.

\gptbox{Candidate A is not correct. The leading $t^{-2}$ term of
$\kappa_3(x,x,x)$ does not vanish. The correct leading asymptotic is
$\kappa_3(x,x,x) \sim -\alpha/\lambda^3 \cdot t^{-2}$. \dots\
You wrote $\langle x^3\rangle_t = -\frac{3\alpha}{2\lambda^3} t^{-2}
+ o(t^{-2})$, but the correct leading term is
$\langle x^3\rangle_t = -\frac{5\alpha}{2\lambda^3} t^{-2} + o(t^{-2})$.}

The lesson generalises: \emph{always run the numerical sanity check
before the GPT consult, even when the Lean machinery is mostly already
in place}. The skill's discipline anticipates this; the Wick coefficient
trap is a concrete instance.

\paragraph{The architectural question.} The candidates survey listed
three options for where the new theorem should live: in threepoint
(re-deriving the anharmonic asymptotics inline), in laplace (with a
local copy of $\kappa_3$ to avoid the cross-repo dependency), or with
laplace adding a \texttt{[[require]]} on threepoint. GPT-5.5~Pro
recommended the third unambiguously:

\gptbox{\textbf{Put the theorem in \texttt{laplace}, and make
\texttt{laplace} depend on \texttt{threepoint} if feasible.} Reason:
\texttt{threepoint} owns the generic cumulant notion. \texttt{laplace}
owns the anharmonic asymptotic machinery. So the natural layering is
generic $\to$ analytic application, i.e.\ \texttt{laplace} imports
\texttt{threepoint}, not the other way around.}

The dependency was added by editing the lakefile\footnote{Add a
\texttt{[[require]]} block with \texttt{name = "Threepoint"},
\texttt{git = "..."}, \texttt{rev = "40398d8"}, then run
\texttt{lake update Threepoint}.} of laplace to point at threepoint at
a fixed commit. The threepoint repository builds in $\sim$30~s on first
import, $\sim$8~s incrementally, and shares laplace's Mathlib v4.29.0
pin. Setup took about three minutes of wall-clock; the build took the
rest.

\paragraph{The $A$-parameter snag.} An early version of the unfolding
lemma left the perturbation observable $A$ abstract\footnote{Concretely,
\texttt{Threepoint.kappa3 volume L A t (fun x => x) (fun x => x)} with
generic $A$.} and tried to prove that this equals
$\langle x^3 \rangle - 3 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3$.
This is wrong: the $A$ parameter is not just the perturbation observable
that drops out at $h = 0$; it also appears as a normal observable
inside terms like $\langle A B \rangle \langle \varphi \rangle$ in the
$\kappa_3$ expansion. The fix was to set $A = (\mathrm{fun}\ x \to x)$
explicitly --- which is what the harmonic instantiation in
threepoint\footnote{\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero}.}
does too; the ``id, id, id'' naming refers to all three slots,
perturbation observable included. With $A = (\mathrm{fun}\ x \to x)$,
the $A$ contributions in the expansion become further $\langle x \rangle$
factors and the polynomial identity goes through.

\paragraph{Rewriting under the integrand lambda.} The unfolding lemma
converts seven perturbed-Gibbs-expectation
terms\footnote{\texttt{Threepoint.gibbsExp~...~0} for each of the seven
$\kappa_3$ summands.} to laplace's \texttt{gibbsExpectation}. After
unfolding the cumulant, the integrand lambdas appear with a
beta-redex,\footnote{Concretely:
\texttt{(fun x => x) w * A w * (fun x => x) w} and similar.} not
already canonicalised to $w$ form. Two failed approaches: rewriting
via the bridge lemma directly\footnote{\texttt{rw} with the bridging
lemma \texttt{threepoint\_gibbsExp\_volume\_zero\_eq}; failed because
the goal had already been auto-converted into the ratio-of-integrals
form.}; and rewriting the integrand shape via a function-extensionality
lemma\footnote{\texttt{rw [h\_cube]} with
\texttt{h\_cube : (fun w => w * w * w) = (fun w => w \^{} 3)}; failed
because the integrand inside the integral was inlined, not in lambda
form.}. The fix: keep the bridge lemma's RHS in
\texttt{gibbsExpectation} form (don't unfold past it), and match the
integrand-lambda shape including the beta-redex literally:

\begin{verbatim}
have h_cube : (fun w => (fun x => x) w * (fun x => x) w
                * (fun x => x) w) = (fun w => w ^ 3) := by
  funext w; ring
\end{verbatim}

This is the kind of friction the laplace \texttt{CLAUDE.md} should
record; added to the followups.

\paragraph{The \texttt{nhds (\ldots)} congruence shuffle.} Three places
in the proof construct a \texttt{Tendsto} via \texttt{Tendsto.mul} or
\texttt{Tendsto.add} and arrive at a limit value like
\texttt{nhds (1/lam + 0)} or \texttt{nhds (2 * C * C * 0)}, which
Lean does \emph{not} simplify to \texttt{nhds (1/lam)} or
\texttt{nhds 0} automatically. The fix is a one-line
\texttt{have h\_lim\_eq : \ldots = \ldots := by ring} followed by
\texttt{rw [h\_lim\_eq] at h\_prod}. Three instances; documented.

\paragraph{Mathlib name gotcha for $1/t \to 0$.} The expected name does
not exist;\footnote{Searched-for: \texttt{tendsto\_one\_div\_atTop\_nhds\_zero}.
Actual: \texttt{tendsto\_inv\_atTop\_zero} (uses $x^{-1}$ rather than
$1/x$).} the bridge is
\verb|by simp only [one_div]; exact tendsto_inv_atTop_zero|.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
\begin{itemize}
\item \texttt{tendsto\_inv\_atTop\_zero}: $1/t \to 0$ as $t \to \infty$
(after a \texttt{one\_div} rewrite).
\item \texttt{Tendsto.mul}, \texttt{Tendsto.add}, \texttt{Tendsto.const\_mul},
\texttt{Tendsto.congr'}, \texttt{tendsto\_const\_nhds}: standard
\texttt{Tendsto}-algebra glue.
\item \texttt{Filter.eventually\_gt\_atTop}: ``eventually $t > 0$''
under \texttt{atTop}, used to safely \texttt{field\_simp} an
$1/t \cdot t = 1$ equality.
\item \texttt{ring}, \texttt{field\_simp}, \texttt{funext}: the usual.
\end{itemize}

\paragraph{Threepoint provided.}
\begin{itemize}
\item \texttt{Threepoint.kappa3}: the abstract third cumulant, defined
generically over a measurable space $(W, \mu)$ and a perturbed Gibbs
expectation $\mathtt{gibbsExp}~\mu~L~A~t~h~\varphi$.
\item \texttt{Threepoint.gibbsExp}: the perturbed Gibbs expectation,
specialised at $h = 0$ to the unperturbed ratio of integrals.
\end{itemize}

\paragraph{Laplace provided.}
The three asymptotics named in \S\ref{sec:strategy}, plus the original
\texttt{anharmonicPotential}, \texttt{gibbsExpectation}, and
\texttt{gibbsCov} definitions.

\paragraph{What was new.}
\begin{itemize}
\item \texttt{threepoint\_gibbsExp\_volume\_zero\_eq}: bridges
$\mathtt{Threepoint.gibbsExp\ volume\ L\ A\ t\ 0\ \varphi}
= \mathtt{Laplace.gibbsExpectation\ L\ t\ \varphi}$. Three lines:
unfold both, simp on the perturbation factor.
\item \texttt{kappa3\_id\_id\_id\_unfold}: unfolds
$\mathtt{kappa3\ volume\ L\ x\ t\ x\ x}$ to
$\langle x^3 \rangle - 3 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3$
in laplace-side moments. Twenty lines including the integrand-lambda
gymnastics.
\item \texttt{kappa3\_id\_id\_id\_eq\_cov\_form}: rewrites the
three-moment expansion in covariance form,
$\mathrm{Cov}[x^2, x] - 2 \langle x^2 \rangle \langle x \rangle + 2 \langle x \rangle^3$.
Pure ring after expanding \texttt{gibbsCov}; eight lines.
\item \texttt{secondMoment\_anharmonic\_asymptotic}: the
$t \cdot \langle x^2 \rangle \to 1/\lambda$ corollary of the existing
mean and variance asymptotics. Forty lines including the proof that
$\langle x \rangle \to 0$.
\item The headline theorem itself: $\sim$80~lines of \texttt{Tendsto}
algebra.
\end{itemize}

The mathematical content is the algebraic identity (which is
elementary) and the limit-value calculation
$-2\alpha/\lambda^3 + \alpha/\lambda^3 = -\alpha/\lambda^3$ (which is
\texttt{ring}). Everything else is plumbing.

\section{Lessons}

\begin{itemize}
\item \textbf{Numerical sanity check first, GPT consult second.} The
Wick coefficient error would have wasted multiple hours of formalisation
if it had survived to Step~3. The 30-line numpy script took two minutes
to write and caught the error decisively. The skill's discipline says
to do the numerical check; the lesson here is that ``before GPT'' is
the right time, not after.

\item \textbf{Cross-repo Tide loops are tractable when the dependency
direction is clean.} Adding \texttt{threepoint} as a Lake dependency
of \texttt{laplace} took three minutes; the build was $\sim$30~s on
first import. The architectural question (where does the theorem live?)
was the only non-trivial decision, and GPT-5.5~Pro resolved it cleanly
by appealing to the generic-to-analytic dependency direction.

\item \textbf{The $A$ parameter in \texttt{Threepoint.kappa3} does not
drop out at $h = 0$.} It also appears as a normal observable inside the
cumulant expansion. ``id, id, id'' includes the $A$ slot, paralleling
\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero}. Worth
landing in \texttt{Threepoint/CLAUDE.md} as a recurring trap.

\item \textbf{The integrand-lambda gymnastics are a recurring laplace
friction.} Three of the four new lemmas in this file ended up doing
\texttt{funext w; ring} rewrites of \texttt{(fun x => x) w * \ldots} to
$w \cdot \ldots$ to canonicalise integrand shapes. Add to
\texttt{laplace/CLAUDE.md}: \emph{when an unfolded
\texttt{gibbsExpectation} has \texttt{(fun x => x) w} beta-redexes in
its integrand, match them literally in the rewrite hypotheses; do not
expect \texttt{ring} to see through them}.

\item \textbf{\texttt{nhds (a + 0)} is not \texttt{nhds a} syntactically.}
The clean idiom is to pre-compute the simplified limit-value as a
\texttt{have h\_lim : a + 0 = a := by ring} and \texttt{rw [h\_lim] at hSum}
before the \texttt{congr'}. Recurs three times in this file. Worth a
generic helper or a tactic note in \texttt{laplace/CLAUDE.md}.

\item \textbf{Numerical sanity scales: a 30-line scipy script doubles
as the closed-form expectation in the tide log.} Recording
$\kappa_3 \cdot t^2 \approx -0.495$ at $t = 256, \lambda = 1, \alpha = 0.5$
in the tide log gives the closed-form expectation; running the same
script at the same parameters during the formalisation gives the
sanity check; the agreement to four significant figures is the
post-formalisation cross-check. Three uses of one artefact.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{Strict-improvement: the $\langle x^3 \rangle_t$ asymptotic
as a corollary.} GPT noted that with the moment expansions exposed,
$t^2 \langle x^3 \rangle_t \to -5\alpha/(2\lambda^3)$ is a $\sim$5-line
corollary of $\mathrm{Cov}[x^2, x] = \langle x^3 \rangle - \langle x^2 \rangle \langle x \rangle$
plus the existing asymptotics. Worth packaging as
\texttt{thirdMoment\_anharmonic\_asymptotic} for downstream consistency
and as a user-facing version of an asymptotic that currently exists
only inside intermediate proofs.

\item \emph{Strict-improvement: the same theorem for affine observables
(scalar-multiple in each slot).} Generalise from $(\varphi, A, B) = (x, x, x)$
to $(b\,x, a\,x, c\,x)$. Multilinearity of $\kappa_3$ extracts the
factor $abc$, and the same parity / asymptotic arguments apply. About
40~lines on top of the present file. Parallels \emph{Candidate A$'$}
from the Tide~5 retrospective.

\item \emph{Next-order asymptotic.} The $t^{-3}$ correction to
$\kappa_3(x, x, x)$ has an explicit constant $C(\lambda, \alpha, \gamma)$
that requires the moment expansions to one higher order. Substantially
larger than the present tide; would benefit from first packaging the
mean and second-moment expansions to one more order in
\texttt{IntegralRemainder.lean}.

\item \emph{Promote the algebraic identity to threepoint.} Currently
\texttt{kappa3\_id\_id\_id\_eq\_cov\_form} is private to laplace. If a
second use materialises (e.g.\ a $\kappa_3$ asymptotic on the
quartic-with-cubic potential, or a different choice of three identical
observables), promote it to a small \texttt{Threepoint/Algebra.lean}
file. The eight-line proof transfers verbatim.

\item \emph{The lean-common promotion candidate.} The
``$\langle x^2 \rangle = \mathrm{Var}[x] + \langle x \rangle^2$ at the
moment-asymptotic level'' construction inside the second-moment helper
is generic --- it works for any potential where the variance and mean
asymptotics are given. Worth a generic helper\footnote{Tentative name:
\texttt{secondMoment\_eq\_var\_plus\_mean\_sq\_tendsto}.} in
\texttt{lean-common} once a second use materialises.

\item \emph{Tag the threepoint paper with \texttt{\textbackslash leanref}.}
The threepoint paper currently has no \texttt{\textbackslash leanref}
tags pointing at this result. Once the next non-trivial threepoint
formalisation lands, do a one-pass tagging consistent with the laplace
primer setup.

\item \emph{Add the new gotchas to \texttt{laplace/CLAUDE.md}.} Three
items: the \texttt{nhds (a + 0)} normalisation pattern, the
integrand-lambda \texttt{(fun x => x) w} matching idiom, and the
``$A$ does not drop out at $h = 0$'' note for cross-repo $\kappa_3$ work.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5~Pro contributed the deliberation-stage advisory at Step~2,
including the independent rederivation of the corrected $5\alpha/2$
Wick coefficient, the architectural recommendation to layer
\texttt{laplace} on top of \texttt{threepoint}, and the cleanest
algebraic-shortcut form of the proof. The full consult and the tide log
itself (with candidate drafts, vote, numerical check, and
result) are preserved in the patterning project's tide-log
directory.\footnote{\texttt{sri/projects/patterning/tide-log/}\allowbreak\texttt{2026-05-06-tide-anharmonic-kappa3.md}
and \texttt{gpt55\_tide\_anharmonic\_kappa3\_v1.md}.}

The seabed for this tide spans the original few-day laplace
formalisation (whose anharmonic covariance theorem and multivariate
machinery were the load-bearing prior
work\footnote{\texttt{cov\_anharmonic\_asymptotic} from
\texttt{retrospective.tex} and the multivariate sharp track from
\texttt{retrospective2.tex}.}) and the threepoint scaffold from Tide~5
(whose harmonic-$\kappa_3$
theorem\footnote{\texttt{kappa3\_harmonic\_id\_id\_id\_eq\_zero}.} gave
the proof template and the ``id, id, id'' naming convention). The
present tide is small only because so much was already in place.

\end{document}
