\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{²}{\ensuremath{^2}}
\newunicodechar{⁴}{\ensuremath{^4}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{√}{\ensuremath{\sqrt{\,}}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≥}{\ensuremath{\geq}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{Bounded-prior quartic with continuous test function}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{6 May 2026}

\begin{document}
\maketitle

\begin{abstract}
A single Tide-loop excursion in the \texttt{laplace} seabed, advancing the
grammar paper precursor sequence by one step: where Tide~6 added the bounded
prior to the pure-quartic partition function, this step adds a Lipschitz
test function $\varphi$ inside the integrand. The
headline\footnote{\texttt{quartic\_lipschitz\_unnormalised\_bounded\_prior}.}
proves $\big|\int_{[-a,a]} \varphi(w)e^{-tw^4/24}\,dw - \varphi(0) Z_a(t)\big|
\le K \cdot \tfrac{1}{2}\sqrt{24\pi/t}$ for $K$-Lipschitz $\varphi$. The new
file adds 211 lines, zero
\texttt{sorry}/\texttt{axiom}/\texttt{native\_decide}; clean Lean
diagnostics on first pass after two small Pi-vs-lambda fixes. The proof
structure follows GPT-5.5 Pro's deliberation-stage tactical advice
(``prove the unnormalised estimate first, derive the normalised version as a
corollary''), which sidesteps a denominator-conditioning subtlety that
trapped my initial \emph{normalised} draft. Run executed in a dedicated
\texttt{git worktree}\,---\,the discipline now mandated by the Tide
skill\,---\,with no concurrent-agent races.
\end{abstract}

\section{Setting}

This Tide step is the seventh excursion outward from the \texttt{laplace}
seabed and the second toward the grammar paper. The earlier
\texttt{tide/grammar-precursor-bounded-prior} excursion (2~May) established
the bounded-prior quartic partition function with an exponentially small
remainder.\footnote{Headline:
$\big|\int_{[-a,a]} e^{-tw^4/24}\,dw - \tfrac{1}{2}(24/t)^{1/4}\Gamma(1/4)\big|
\le \tfrac{24}{ta^3} e^{-ta^4/24}$, in
\texttt{Laplace.OneD.QuarticBoundedPrior}.}
That excursion's Follow-up section explicitly named the present candidate as
its natural successor: continuous (or Lipschitz) test functions $\varphi$
against the same bounded-prior quartic Gibbs measure.

The Step~2 deliberation\,---\,which started by drafting three candidate
scopings\,---\,landed on a sharpening that wasn't on Claude's initial draft:
prove the \emph{unnormalised} Lipschitz estimate
$|{\int \varphi g - \varphi(0)\!\int g}| \le K\,\tfrac{1}{2}\sqrt{24\pi/t}$
first, then derive the normalised
$|\langle\varphi\rangle_{t,a} - \varphi(0)| = O(t^{-1/4})$ as a corollary.
This is GPT's contribution and was the right call (see \S\emph{Roadblocks}
below for why). The full deliberation, with the GPT consult preserved
verbatim, lives in the primer project's
tide-log.\footnote{\texttt{projects/primer/tide-log/}\allowbreak\texttt{2026-05-06-tide-bounded-prior-continuous-test.md}.}

The branch is chained on top of the \texttt{tide/separable-potential} tip,
itself one of the two unmerged tides at the start of this excursion (the
other was \texttt{tide/cross-susc-deriv-harmonic}, on a parallel and
independent piece of laplace--threepoint cross-work). The linear-chain
discipline\,---\,each new Tide branched off the previous one's tip rather
than off \texttt{main}\,---\,keeps the squash-merge story straightforward.

\section{The thing formalised}

For the centred quartic potential $L(w) = w^4/24$, any $a, t > 0$, $K \ge 0$,
and any continuous $\varphi : \mathbb{R} \to \mathbb{R}$ satisfying the
Lipschitz-near-zero condition
\[
  |\varphi(w) - \varphi(0)| \le K \cdot |w|
  \qquad \text{for all } w \in [-a, a],
\]
the bounded-prior unnormalised numerator differs from $\varphi(0) Z_a(t)$ by
at most a closed-form constant times $t^{-1/2}$:

\begin{quote}\itshape
\[
  \left|\int_{[-a,a]} \varphi(w)\,e^{-tw^4/24}\,dw
        \;-\; \varphi(0) \int_{[-a,a]} e^{-tw^4/24}\,dw\right|
  \;\le\; K \cdot \tfrac{1}{2} \cdot \sqrt{24\pi / t}.
\]
\end{quote}

\noindent The Lean signature:
\begin{verbatim}
theorem quartic_lipschitz_unnormalised_bounded_prior
    {t a K : ℝ} (ht : 0 < t) (ha : 0 < a) (hK : 0 ≤ K)
    {φ : ℝ → ℝ} (hφ : Continuous φ)
    (hLip : ∀ w ∈ Icc (-a) a, |φ w - φ 0| ≤ K * |w|) :
    |(∫ w in Icc (-a) a, φ w * exp (-(t * w ^ 4 / 24))) -
        φ 0 * (∫ w in Icc (-a) a, exp (-(t * w ^ 4 / 24)))| ≤
      K * (1/2) * Real.sqrt (24 * Real.pi / t)
\end{verbatim}

\noindent located in \texttt{Laplace.OneD}.

The grammar paper's $Z_n[\varphi]$ machinery wants three ingredients on top
of the existing seabed: (i) bounded prior $\pi$, (ii) test function
$\varphi$ inside the integrand, (iii) explicit asymptotic in $n$ with a
remainder. Tide~6 added (i); the present tide adds (ii). What remains is
the explicit $o(t^{-1/4})$ packaging of (iii), now reachable as a follow-up
that combines this tide's bound with the existing
\texttt{quartic\_partition\_bounded\_prior} (an exponentially small
correction to $Z_a(t)$).

The unnormalised rate is $t^{-1/2}$ in this statement; the normalised
$|\langle\varphi\rangle_{t,a} - \varphi(0)|$ then gets the
``substitution-driven'' rate $t^{-1/4}$ once one divides by
$Z_a(t) \sim Z_{\text{full}}(t) = O(t^{-1/4})$. The closed-form rate
constant for the normalised version is
$C = 24^{1/4}\sqrt{\pi}/\Gamma(1/4) \approx 1.082$, but that is a corollary,
not the headline.

Numerical sanity check (run before formalisation):
\texttt{scipy.integrate.quad} verified all three deliberation-stage
candidates' closed forms to machine precision across five values of $t$
ranging from $10$ to $10^5$, with several test
functions.\footnote{$\varphi \in \{1, w, w^2, |w|, \cos w, \cos w +
\tfrac{1}{2}\}$.} The $\beta$-rate constant $C$ matched the predicted
value of $1.082048$ to machine precision for $\varphi = |w|$ at
$t \ge 1000$.

\section{Proof strategy}

The unnormalised statement reduces, via the trivial identity
\[
  \int \varphi g - \varphi(0) \int g \;=\; \int (\varphi - \varphi(0)) g,
\]
to bounding $|\int (\varphi - \varphi(0)) g|$. Pointwise on $[-a, a]$,
$|(\varphi - \varphi(0)) g| \le K \cdot |w| \cdot g$ by the Lipschitz
hypothesis (the Boltzmann factor $g(w) = e^{-tw^4/24}$ is positive, so its
absolute value is itself). The triangle-for-integrals lemma
$|\int f| \le \int |f|$ then takes the absolute value inside, and the
pointwise bound followed by monotonicity of the integral
(\texttt{setIntegral\_mono\_on}) reduces the whole thing to bounding
$\int_{[-a,a]} |w|\,g(w)\,dw$.

The bounded-prior first absolute moment is bounded by its full-line
counterpart by monotonicity over a subset (the integrand is
non-negative). The full-line first absolute moment has a clean closed
form:
\[
  \int_{\mathbb{R}} |w|\,e^{-tw^4/24}\,dw \;=\; \tfrac{1}{2}\sqrt{24\pi/t}.
\]
This is the substantive new computation in this tide. The proof goes
through the half-line case
$\int_{(0,\infty)} w\,e^{-tw^4/24}\,dw = \tfrac{1}{4}\sqrt{24\pi/t}$, which
falls directly out of Mathlib's master gamma-integral
lemma\footnote{\texttt{integral\_rpow\_mul\_exp\_neg\_mul\_rpow}.} at
$p = 4$, $q = 1$, $b = t/24$, with the closed-form value
$(t/24)^{-1/2}\,\tfrac{1}{4}\,\Gamma(1/2)$, which simplifies to
$\tfrac{1}{4}\sqrt{24\pi/t}$ (since $\Gamma(1/2) = \sqrt{\pi}$ and
$(t/24)^{-1/2} = \sqrt{24/t}$). The full-line case is twice the half-line
case by parity of the integrand, via Mathlib's
\texttt{integral\_comp\_abs}\,---\,exactly the same routing established by
the Quartic and Sextic moment files.

So the chain is:

\begin{enumerate}\itemsep0pt
\item Half-line first moment, closed form (master lemma).
\item Full-line first absolute moment, double of (1) by parity.
\item Bounded-prior first absolute moment $\le$ full-line, by set monotonicity.
\item Apply Lipschitz pointwise, triangle-inequality, set monotonicity, and (3).
\end{enumerate}

Each link is a single Mathlib invocation modulo bookkeeping; the file is
211 lines because the bookkeeping (rpow-vs-npow casts, Pi.sub-vs-lambda
mismatches, integrability witnesses) is unavoidable at present.

\section{Roadblocks and resolutions}

\paragraph{The deliberation-stage pivot from normalised to unnormalised.}
Claude's initial Step~2 draft proposed the normalised statement
$|\langle\varphi\rangle_{t,a} - \varphi(0)| \le KC t^{-1/4}$ as the
target. GPT-5.5 Pro pointed out that proving this directly trips on a
subtle conditioning identity:

\gptbox{The truncated Gibbs law on $[-a,a]$ is the full-line quartic law
conditioned on $|W|\le a$, so its $|W|$-expectation is $\le$ the full-line
$|W|$-expectation. That full-line expectation is exactly $C t^{-1/4}$.}

That paragraph reads cleanly, but \emph{it is not quite right}: for a
non-negative random variable $Y$ and an event $A$, the conditional
expectation $E[Y \mid A] = E[Y \cdot \mathbf{1}_A]/P(A)$, which is bounded
\emph{above} by $E[Y]/P(A)$, not by $E[Y]$. For our problem
$P(|W| \le a) \to 1$ exponentially as $t \to \infty$, so
$1/P(|W| \le a) = 1 + (\text{exp-small})$ and the normalised bound
recovers an exp-small correction GPT's first-pass argument suppresses.

The unnormalised statement, by contrast, dodges all conditioning: the
inequality
\[
  \int_{[-a,a]} |w|\,g(w)\,dw \;\le\; \int_{\mathbb{R}} |w|\,g(w)\,dw
\]
follows from set monotonicity of the integral with no probability
arithmetic at all. GPT's broader recommendation\,---\,``prove the
unnormalised version first, then derive the normalised one as a
corollary''\,---\,is exactly right, and it's the right architectural choice
even if the specific justification needs a small correction. The
unnormalised version landed cleanly here; the normalised corollary now
becomes a separate, smaller follow-up tide that can carry the partition
ratio analysis explicitly.

\paragraph{Two Pi-vs-lambda mismatches.} Both familiar from the project's
\texttt{CLAUDE.md} cookbook. The first appeared in the integrability of
$|w|\cdot g(w)$: \texttt{Integrable.abs} returns the integrability of
$|w \cdot g(w)|$, but the headline wants $|w| \cdot g(w)$, and converting
between them needs $|w \cdot g| = |w| \cdot |g| = |w| \cdot g$ (since
$g > 0$). The fix is the standard
$\mathtt{abs\_mul} \to \mathtt{abs\_of\_pos\ (Real.exp\_pos\ \_)}$ rewrite
inside a $\mathtt{funext}$. The second appeared in subtracting two
integrable functions on $[-a,a]$: \texttt{IntegrableOn.sub} returns
\texttt{IntegrableOn (f - g) ...} with $f - g$ as $\mathtt{Pi.sub}$, but
the headline expects the single-lambda
$\mathtt{fun\ w\ =>\ (\varphi\ w - \varphi\ 0)\cdot g\ w}$. The fix is to
prove a $\mathtt{funext}$ equality and \texttt{rwa}; same recipe as the
quartic-cov-affine and pure-quartic tides. Both fixes were one-shot.

\paragraph{The rpow-to-sqrt conversion.} Going from $(t/24)^{-1/2}$ to
$\sqrt{24/t}$ took a few attempts through different Mathlib idioms before
landing on a four-step rewrite chain.\footnote{\texttt{Real.rpow\_neg}
$\to$ \texttt{Real.sqrt\_eq\_rpow} $\to$ \texttt{Real.sqrt\_inv} $\to$
\texttt{congr 1; field\_simp}.} This sequence is not in the project's
\texttt{CLAUDE.md} yet; should be added.

\paragraph{Worktree isolation worked as designed.} The Tide skill now
mandates running each Tide in its own \texttt{git worktree}. This run was
the first to use that discipline from the start: the worktree at
\texttt{/tmp/laplace-bounded-test/} was created before any work began, and
no concurrent-agent races materialised. Compare with the earlier
2~May Tides where multiple concurrent agents racing on
\texttt{Laplace.lean} cost an extra round-trip per Tide. The lesson is now
load-bearing in the skill, not a footnote.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
The master half-line gamma integral; the even-symmetric reduction to the
half line; the closed-form $\Gamma(1/2) = \sqrt{\pi}$; the triangle
inequality $|\int f| \le \int |f|$; pointwise-bound monotonicity; set
monotonicity (subset $\mathtt{Icc}$ into all of $\mathbb{R}$); the
rpow-to-sqrt conversion lemmas; the integration-by-parts lemmas for linear
combinations.\footnote{Specifically:
\texttt{integral\_rpow\_mul\_exp\_neg\_mul\_rpow},
\texttt{integral\_comp\_abs},
\texttt{Real.Gamma\_one\_half\_eq},
\texttt{MeasureTheory.abs\_integral\_le\_integral\_abs},
\texttt{MeasureTheory.setIntegral\_mono\_on},
\texttt{MeasureTheory.setIntegral\_le\_integral},
\texttt{Real.rpow\_neg}, \texttt{Real.sqrt\_eq\_rpow},
\texttt{Real.sqrt\_inv}.}

\paragraph{The seabed provided.}
\texttt{quartic\_integrable\_pow} from \texttt{Laplace.OneD.Quartic} (for
the full-line integrability of $w^n e^{-tw^4/24}$). The tide's import is
just \texttt{Laplace.OneD.QuarticBoundedPrior}; the partition function and
\texttt{quartic\_integrable\_pow} chain transitively from there.

\paragraph{What was new in this tide.}
The half-line first moment, the full-line first absolute moment, the
integrability witness, the bounded-prior comparison, and the headline
unnormalised Lipschitz estimate.\footnote{Lean names:
\texttt{quartic\_integral\_w\_exp\_Ioi},
\texttt{quartic\_integral\_abs\_w\_exp\_full},
\texttt{quartic\_integrable\_abs\_w},
\texttt{quartic\_integral\_abs\_w\_bounded\_prior\_le},
\texttt{quartic\_lipschitz\_unnormalised\_bounded\_prior}.}
Five new theorems; the four sublemmas are general enough they could be
promoted to \texttt{lean-common} on the next bounded-prior tide that needs
them.

\section{Lessons}

\begin{itemize}
\item \emph{Unnormalised before normalised.} The Step~2 deliberation taught
the right architectural lesson for grammar-paper-style asymptotics: when
the headline statement is a normalised ratio
$\langle\varphi\rangle_{t,a} \to \varphi(0)$, the cleanest formalisation
path goes through the unnormalised numerator difference first. The
normalisation\,---\,dividing by the partition function\,---\,brings in
denominator gymnastics (lower bound on $Z_a$, ratio of two limits) that
are easier to handle once the unnormalised piece exists as a stand-alone
theorem. Worth lifting to the Tide skill or to
\texttt{lean-formalisation}'s playbook.

\item \emph{Trust GPT's strategic recommendations even when the
justification has a small flaw.} GPT's ``the truncated Gibbs is conditioned
on $|W| \le a$, so its $|W|$-expectation is $\le$ the full-line one'' was
not quite right (it's bounded by $E[|W|]/P(|W| \le a)$, with a $1/P$
factor that is benign but not invisible). But the broader architectural
recommendation\,---\,prove unnormalised first\,---\,was completely right,
and the right mental move was to take the architectural advice and
substitute the cleaner \emph{set-monotonicity} justification for the
sketch GPT gave.

\item \emph{The rpow-to-sqrt cast chain.} The four-step rewrite\footnote{Same
sequence as above: \texttt{Real.rpow\_neg} $\to$ \texttt{Real.sqrt\_eq\_rpow}
$\to$ \texttt{Real.sqrt\_inv} $\to$ \texttt{congr 1; field\_simp}.} recurs
in any Tide that converts a half-line $\Gamma(1/2)$ result to its sqrt
presentation. Worth a one-line entry in \texttt{laplace}'s
\texttt{CLAUDE.md} so the next session doesn't re-derive it.

\item \emph{Worktree isolation is now load-bearing.} The Tide skill's
preflight rule (every Tide in its own \texttt{git worktree}) was followed
from the start and the run was free of branch-toggle races. This is the
first run that hit zero concurrent-agent friction. The discipline is paying
for itself.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{Normalised version: $|\langle\varphi\rangle_{t,a} - \varphi(0)|
\le KC\,t^{-1/4} + (\text{exp-small})$.} Direct corollary combining the
present unnormalised estimate with the existing
\texttt{quartic\_partition\_bounded\_prior}: lower-bound $Z_a(t)$ by
$(1/2)(24/t)^{1/4}\Gamma(1/4) - \tfrac{24}{ta^3}e^{-ta^4/24}$, divide. The
explicit constant $C = 24^{1/4}\sqrt{\pi}/\Gamma(1/4)$ is the rate
constant; numerically $C \approx 1.082048$. A quick follow-up tide
($\sim$60--100 lines).

\item \emph{$\mathtt{IsLittleO}$ packaging.} For continuous $\varphi$
(weaker than Lipschitz), the limit
$\langle\varphi\rangle_{t,a} \to \varphi(0)$ holds qualitatively\,---\,the
rate $t^{-1/4}$ is special to Lipschitz $\varphi$\,---\,and Mathlib's
asymptotic-equivalence API\footnote{\texttt{Asymptotics.IsLittleO}.} would
be the form needed for the grammar paper's
$Z_n[\varphi] \sim C_\varphi n^{-\lambda} (\log n)^{m-1}$. This is a
separate, larger Tide.

\item \emph{Bounded-prior monomial generalisation $L = w^{2k}/(2k)!$.}
The full chain transfers verbatim to the sextic ($k=3$), octic ($k=4$),
and so on. Pairs naturally with the morning survey's A1 candidate
(``generic even-monomial 1D template''), which is now strict-improvement
territory: with three $k$-instances completed (quartic, sextic, and now
the bounded-prior version of the quartic), the case for a unifying
parameterised file is overwhelming.

\item \emph{Promote the splitting/union and abs-of-product idioms.}
Several small lemmas in this file ($\mathtt{abs\_mul}$ +
$\mathtt{abs\_of\_pos}$ for an exp-positive Boltzmann factor; the
$|w \cdot g| = |w| \cdot g$ rewrite chain; the unnormalised-difference =
$\int (\varphi - \varphi(0)) g$ identity) recur on every Tide that
involves a Lipschitz observable against a Gibbs measure. Worth promoting
to \texttt{lean-common} on the next use.

\item \emph{Continuous (not Lipschitz) test function via
modulus-of-continuity.} GPT's missed candidate (i): replace the
Lipschitz-rate hypothesis with a modulus-of-continuity statement. The
qualitative limit ($\alpha$) becomes a clean $\varepsilon/\delta$
argument; smoothing parameter $\delta$ chosen via the tail bound. Cleaner
formulation for ``continuous test functions'' than the Lipschitz cut, but
larger formalisation.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step~2,
including the architectural recommendation to prove the unnormalised
estimate first (the recommendation that landed) and the
sharpened-but-slightly-imprecise justification for it (the conditioning
argument that needed correcting on the way to the proof). The full consult
and the tide log itself are preserved together in the primer project's
tide-log directory.\footnote{\texttt{projects/primer/tide-log/}\allowbreak\texttt{gpt55\_tide\_bounded\_prior\_continuous\_test\_v1.md}
and the markdown tide log alongside it.}

\end{document}
