\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{²}{\ensuremath{^2}}
\newunicodechar{⁴}{\ensuremath{^4}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{λ}{\ensuremath{\lambda}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{2D pure-quartic affine covariance}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{2 May 2026}

% Allow LaTeX a bit of extra interword stretch for paragraphs that
% contain long unbreakable identifiers (e.g. Lean theorem names like
% \texttt{cov_affine_pureQuartic}). 3em is enough for the residual
% ~36-40pt overfulls that survive even after footnoting the longest
% identifiers.
\emergencystretch=3em

\begin{document}
\maketitle

\begin{abstract}
A single Tide-loop excursion in the \texttt{laplace} seabed, formalising the
affine covariance for the doubly-degenerate 2D potential
$L(x,y) = (x^4 + y^4)/24$. The headline theorem
\texttt{cov\_affine\_pureQuartic}\footnote{In the new file
\texttt{Laplace/TwoD/PureQuartic.lean}, namespace
\texttt{Laplace.TwoD.PureQuartic}.} adds 478 lines of Lean to the
seabed; the proof is a near-mechanical mirror of the existing
\texttt{cov\_affine\_semiDegenerate}, with the harmonic $y$-direction
replaced by a second copy of the quartic. Zero \texttt{sorry}, zero
\texttt{axiom}, zero \texttt{native\_decide}; clean Lean diagnostics on
first pass after a namespace fix. The interesting friction in this run was
not mathematical: two other Tide agents (sextic, bounded-prior) were
operating in the same checkout concurrently, and one of them clobbered
\texttt{Laplace.lean} between \texttt{git add} and \texttt{git commit}.
The fix was a trailing single-line commit; the lesson is to run
concurrent Tide agents in separate \texttt{git worktree} checkouts.
\end{abstract}

\section{Setting}

This Tide step is the fifth excursion outward from the same seabed: the
\texttt{laplace} repo, which originally formalised the 1D anharmonic
asymptotic
$\operatorname{Cov}_t[x^2,x] = -2\alpha/(\lambda^3 t^2) + o(t^{-2})$
(Susceptibility Primer, eq.~4.10) and has since been extended by four
prior tide steps\footnote{Branches: \texttt{tide/quartic-moments} (1D
pure-quartic case), \texttt{tide/2d-semi-degenerate} (2D semi-degenerate
case), \texttt{tide/universal-scaling} (universal scaling identity), and
\texttt{tide/f1-large-t} (large-$t$ asymptotics of $f_1$). All four were
merged into \texttt{main} on 1\,May.}. This step branches off the
resulting \texttt{main}, commit \texttt{44a6001}.

The candidate was picked from a cross-tide candidate
survey,\footnote{\texttt{projects/automation/log/2026-05-02-tide-candidates-survey.md}}
which catalogued five candidate steps drawn from the wider research
portfolio. Of those, the 2D pure-quartic case scored as the cleanest
mechanical extension: the existing semi-degenerate 2D
file\footnote{\texttt{Laplace/TwoD/SemiDegenerate.lean} (existing 2D
seabed) and \texttt{Laplace/OneD/Quartic.lean} (1D quartic seabed).}
already provides potential-agnostic 2D wrappers
(\texttt{partitionFunction}, \texttt{gibbsExpectation},
\texttt{gibbsCov}) and a worked example of the $\exp(-tL)$
factorisation pattern; the existing 1D quartic file provides closed-form
1D moments. Combining them is a substitution exercise.

The Step~2 deliberation (Claude $\leftrightarrow$ GPT-5.5 Pro) drafted
three candidates: A, the affine covariance itself; B, the more general
mixed even-power moment family $\langle x^{2m} y^{2n}\rangle_t$; C, just
the moment table $\langle z_j\rangle$, $\langle z_1 z_2\rangle$,
$\langle z_j^2\rangle$. Both parties backed A on the same grounds: A is
the smallest statement that closes a useful interface (the affine
covariance is what subsequent susceptibility work will call), and B's
closed form is mathematically clean but introduces \texttt{Real.rpow}
exponent bookkeeping (half-integral when $m+n$ is odd) that costs more
in Lean than its incremental power justifies. The full deliberation is
in the tide
log.\footnote{\texttt{projects/primer/tide-log/2026-05-02-tide-2d-pure-quartic.md}}

\section{The thing formalised}

For the centred quartic-quartic 2D potential
\[
  L(x,y) = \frac{x^4}{24} + \frac{y^4}{24}
\]
and affine observables $\varphi(x,y) = a_1 x + a_2 y + c$,
$\psi(x,y) = b_1 x + b_2 y + d$, against the Gibbs measure
$e^{-t L(x,y)}\,dx\,dy / Z(t)$ on $\mathbb{R}^2$:

\begin{quote}\itshape
For all $t > 0$,
\[
  \operatorname{Cov}_t[\varphi,\psi]
  \;=\; (a_1 b_1 + a_2 b_2) \cdot \sqrt{24/t} \cdot
        \frac{\Gamma(3/4)}{\Gamma(1/4)}.
\]
\end{quote}

\noindent The Lean signature:
\begin{verbatim}
theorem cov_affine_pureQuartic {t : ℝ} (ht : 0 < t)
    (a₁ a₂ b₁ b₂ c d : ℝ) :
    gibbsCov pureQuarticPotential t
        (fun z : ℝ × ℝ => a₁ * z.1 + a₂ * z.2 + c)
        (fun z : ℝ × ℝ => b₁ * z.1 + b₂ * z.2 + d) =
      (a₁ * b₁ + a₂ * b₂) * Real.sqrt (24 / t) *
        Real.Gamma ((3 : ℝ) / 4) / Real.Gamma ((1 : ℝ) / 4)
\end{verbatim}

\noindent in the new file \texttt{Laplace/TwoD/PureQuartic.lean}, namespace
\texttt{Laplace.TwoD.PureQuartic}.

The statement is the doubly-degenerate analogue of the existing
\texttt{cov\_affine\_semiDegenerate}: where the latter has one harmonic
direction (rate $t^{-1}$) and one quartic direction (rate $t^{-1/2}$), the
present theorem has both directions degenerate and a single rate
$t^{-1/2}$. The constants $c, d$ drop out (covariance is shift-invariant in
each argument); the cross terms $a_1 b_2$ and $a_2 b_1$ vanish because
$\langle z_1 z_2\rangle = 0$ by parity in either factor; what survives is
the diagonal contribution $(a_1 b_1 + a_2 b_2) \langle z_j^2\rangle$, with
$\langle z_j^2\rangle = \sqrt{24/t}\,\Gamma(3/4)/\Gamma(1/4)$ from the 1D
quartic moment lemma.

Numerical sanity check: \texttt{scipy.integrate.dblquad} over $[-6,6]^2$
agreed with the closed form to $1.4\times 10^{-14}$ across four
$(a_i,b_j,c,d,t)$ test points (and analogously to machine precision for
candidates B and C; all are recorded in the tide log).

\section{Proof strategy}

The Boltzmann factor $\exp(-t L(x,y))$ separates as
\[
  \exp(-t (x^4 + y^4)/24)
  = \exp(-t \cdot x^4/24) \cdot \exp(-t \cdot y^4/24),
\]
so the 2D Gibbs measure is the product of two copies of the 1D quartic
Gibbs measure. The proof rides this factorisation throughout. The
load-bearing Mathlib lemma is
\texttt{integral\_prod\_mul}\footnote{In the
\texttt{MeasureTheory} namespace.}: under no hypotheses, the integral
over $\mathbb{R}^2$ of a separable function $f(x)\,g(y)$ equals the
product of the 1D integrals (and is $0$ on both sides if non-integrable,
by Bochner convention). Applied to the partition function this gives
\texttt{partitionFunction\_factor}: $Z_{2D}(t) =
Z_{\text{quartic}}(t)^2$. Applied to monomials it gives
\texttt{integral\_pow\_pow\_factor}:
\[
  \int_{\mathbb{R}^2} z_1^m z_2^n e^{-t L(z)} \,dz
  \;=\; \Big(\int_\mathbb{R} x^m e^{-t x^4/24}\,dx\Big)
        \Big(\int_\mathbb{R} y^n e^{-t y^4/24}\,dy\Big).
\]
The 1D factors are then \texttt{quartic\_moment\_even} and
\texttt{quartic\_moment\_odd} (zero by parity).

From there the proof is mechanical: the means $\langle z_j\rangle = 0$
and the mixed first moment $\langle z_1 z_2\rangle = 0$ vanish via the
parity factor; the second moments $\langle z_j^2\rangle$ reduce via
\texttt{quartic\_expected\_value\_sq}; the headline covariance follows by
expanding $(a_1 z_1 + a_2 z_2 + c)(b_1 z_1 + b_2 z_2 + d)$ and using
linearity of \texttt{gibbsExpectation} (which itself unfolds against
\texttt{integral\_add} and \texttt{integral\_const\_mul}, having first
established 2D atom integrability for each $z_1^m z_2^n e^{-t L}$ via
\texttt{Integrable.mul\_prod} from two
\texttt{quartic\_integrable\_pow\_pot} witnesses).

The proof body of \texttt{cov\_affine\_pureQuartic} is a six-term
\texttt{integral\_add} cascade (six monomial cross-terms in the
expansion of the product of the two affine observables) followed by a
\texttt{field\_simp} and \texttt{ring} closer. This is verbatim the same
skeleton as \texttt{cov\_affine\_semiDegenerate}; the only changes are
the potential, the namespaces, and the absence of the harmonic-direction
integrability helper.

\section{Roadblocks and resolutions}

\paragraph{Namespace clash with the existing 2D excursion.}
The first build attempt produced a wall of \emph{has already been
declared} errors. The cause was straightforward: the existing
semi-degenerate file put its helper lemmas\footnote{For example,
\texttt{partitionFunction\_factor} and
\texttt{gibbsExpectation\_fst\_sq}.} at unqualified names inside
\texttt{namespace Laplace.TwoD}\,---\,under the same generic names that
the present file naturally wants for the quartic-quartic potential.
With both files in the same namespace, the lemma-name space collides.

The cheap fix (taken here) is to put the present file's lemmas in a
nested sub-namespace \texttt{Laplace.TwoD.PureQuartic}; the bare
identifiers \texttt{partitionFunction}, \texttt{gibbsExpectation}, and
\texttt{gibbsCov} still resolve because Lean searches outward for
unqualified names. This is the lighter of the two routes.

The heavier and structurally cleaner route, suggested by GPT-5.5 Pro at
the deliberation step and deferred here, is to extract a generic
\emph{additively-separable potential} factorisation: prove once that for
any $L(x,y) = U(x) + V(y)$ the partition function factorises and
expectations of separable observables factorise, then specialise to
$U = V = x^4/24$ (this file) and to $U = x^4/24,\,V = \lambda y^2/2$
(the existing semi-degenerate file). With this abstraction in place,
GPT noted, ``B becomes almost free''\,---\,i.e.\ the mixed even-power
moment table $\langle x^{2m} y^{2n}\rangle$ falls out without further
work. The decision to defer was deliberate: the Tide skill prefers
small finished steps to medium-sized refactors, and the namespace
pressure is manageable at two 2D potentials. It will not be at three.

\paragraph{Concurrent-agent staging race.}
The repository hosting this Tide step had two other Tide branches
running simultaneously\footnote{The branches were
\texttt{tide/sextic-moments} (a 1D sextic excursion) and
\texttt{tide/grammar-precursor-bounded-prior} (an in-progress excursion
toward the grammar paper).}, all operating in the same checkout.
Multiple times during this run, the working-tree copy of
\texttt{Laplace.lean} was overwritten by another agent who was also
adding an import for their new file. Once, the overwrite happened in
the window between \texttt{git add Laplace.lean} and
\texttt{git commit}\,---\,the staged content survived, but Git's index
appears to have been touched again before the commit fired, with the
result that the new file's \texttt{import} line did not make it into
the first commit (\texttt{1e8954b}). The working tree then showed
\texttt{Laplace.lean} as unmodified relative to the new \texttt{HEAD},
so a quick \texttt{git diff} did not surface the problem; only
inspecting the commit's file list revealed it.

The fix was a single-line trailing commit (\texttt{efafbfa}) wiring the
import. This worked but is not a satisfying mitigation. The operational
lesson, written up under \emph{Lessons} below, is that concurrent Tide
agents should run in separate \texttt{git worktree} checkouts so that
their working-tree edits cannot interleave.

Beyond these two, no proof-side roadblocks: the proof typechecked clean
on the first compile after the namespace fix, with zero warnings. The
candidate had been picked precisely so that the seabed's machinery would
already cover the moves; the run paid the cost of that prediction.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
\texttt{integral\_prod\_mul} for the partition-function and moment
factorisations. \texttt{Integrable.mul\_prod} for promoting two 1D
integrability witnesses to a 2D one. \texttt{Real.exp\_add} for the
Boltzmann factor decomposition. The standard Bochner integral lemmas
\texttt{integral\_add}, \texttt{integral\_const\_mul}, and
\texttt{integral\_neg} for the affine-observable expansion.
\texttt{mul\_div\_mul\_left} and \texttt{mul\_div\_mul\_right} for the
$Z_{\text{quartic}}$-cancellation in the second-moment lemmas.

\paragraph{The seabed provided.}
This step did not consume new Mathlib API; everything not shipped by
Mathlib was already in the seabed. From the 1D quartic file, the
partition function and its positivity, the even and odd full-line
moment integrals, the second-moment expected value, and a polynomial
integrability witness against the quartic Gibbs
weight.\footnote{Specifically: \texttt{quartic\_partition},
\texttt{quartic\_partition\_pos}, \texttt{quartic\_moment\_even},
\texttt{quartic\_moment\_odd}, \texttt{quartic\_expected\_value\_sq},
and \texttt{quartic\_integrable\_pow\_pot}.} From the existing 2D
semi-degenerate file, the potential-agnostic
\texttt{partitionFunction}, \texttt{gibbsExpectation}, and
\texttt{gibbsCov} wrappers (which this file reuses unchanged) and the
proof skeleton that this file mirrors.

\paragraph{What was new in this tide.}
The 2D pure-quartic potential
\texttt{pureQuarticPotential}, its Boltzmann factor decomposition, five
moment lemmas (the two means, the mixed first moment, and the two
second moments\footnote{Named \texttt{gibbsExpectation\_X} for
\texttt{X} $\in \{$\texttt{fst}, \texttt{snd}, \texttt{fst\_mul\_snd},
\texttt{fst\_sq}, \texttt{snd\_sq}$\}$. The first-moment lemmas are
visibly symmetric, but Lean wants both because of the asymmetric
structure of the underlying product space.}),
a private 2D atom integrability witness, and the headline
\texttt{cov\_affine\_pureQuartic}. There is no novel mathematical
content; what is new is the substitution-into-the-template that gives a
working Lean proof.

\section{Lessons}

\begin{itemize}
\item \emph{Concurrent Tide agents need filesystem isolation.} The
\texttt{Laplace.lean} import-list is a contention point because every
new top-level file edits it. Two agents in the same checkout will race;
it is not a question of careful sequencing. The structural fix is to
run each Tide agent in its own \texttt{git worktree} attached to its
own branch. Worth lifting to the Tide skill as a preflight step
alongside the existing one-tide-branch-per-repo rule.

\item \emph{Generic lemma names in the master 2D namespace are a
smell.} The existing partition-function factorisation lemma reads, by
its short name, as if it were a property of the
namespace\,---\,but it is in fact specific to one potential. A second
potential can only coexist by sub-namespacing or by suffixing every
lemma; neither scales. The right time to extract a generic
\emph{additively-separable factorisation} is now (two potentials), not
later (three or four). This is the deferred-but-already-cost-positive
refactor noted in the follow-ups.

\item \emph{The substitution template is robust at one more
exponent.} This step confirmed that the
$\exp(-tL_1) \cdot \exp(-tL_2)$ factorisation ports cleanly across the
quartic--quartic and harmonic--quartic boundary. The next test is
whether it ports across the sextic side, courtesy of the concurrent
sextic Tide step\footnote{The 1D sextic file
\texttt{Laplace/OneD/Sextic.lean} from
\texttt{tide/sextic-moments}.}; if so, the abstraction is genuinely
the right shape.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{Generic separable-potential factorisation.} For any
$L(x,y) = U(x) + V(y)$, a single file proving the partition-function
factorisation, the separable-observable factorisation
$\langle f(x) g(y)\rangle = \langle f\rangle_U \langle g\rangle_V$,
and optionally the bilinear-symmetric-affine-shift algebra of
\texttt{gibbsCov}. Specialises to both the semi-degenerate and
pure-quartic cases; candidate B (mixed even-power 2D moments) drops
out for free. Best opened as a cleanup PR after the next 2D Tide step
lands.

\item \emph{Candidate B harvesting.} The mixed even-power moment table
$\langle x^{2m} y^{2n}\rangle_t = (24/t)^{(m+n)/2} \cdot \Gamma((2m+1)/4)
\Gamma((2n+1)/4)/\Gamma(1/4)^2$ is now reachable in $\sim$\,30 lines from
the present file (or in $\sim$\,3 lines from the separable-potential
abstraction above). A small future Tide could harvest it; it is more
naturally a corollary of the abstraction than a standalone target.

\item \emph{Quartic--sextic mixed potential.} $L(x,y) = x^4/24 +
y^6/720$, with rates $t^{-1/2}$ and $t^{-1/3}$ respectively. Mechanical
given the present file plus the concurrent sextic seabed. Worth a Tide
step once that file merges.

\item \emph{Worktree-isolated Tide runs.} Process change rather than
maths: prepare a small wrapper that creates a \texttt{git worktree}
for each Tide branch and runs the Tide agent against that worktree.
Avoids the staging race observed in this run.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step~2,
including the \texttt{Real.rpow}-vs-\texttt{Nat}-pow caveat for
candidate B and the deferred separable-potential refactor suggestion.
The full consult and the tide log itself (with candidate drafts, vote,
numerical check, and result) are preserved in the primer
tide-log.\footnote{GPT response:
\texttt{projects/primer/tide-log/gpt55\_tide\_2d\_pure\_quartic\_v1.md}.
Tide log:
\texttt{projects/primer/tide-log/2026-05-02-tide-2d-pure-quartic.md}.}

\end{document}
