\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{κ}{\ensuremath{\kappa}}

\title{Tide retrospective:\\affine-observable $\kappa_3$ for the anharmonic Gibbs (G2)}
\author{Daniel Murfet}
\date{7 May 2026}

\begin{document}
\maketitle

\begin{abstract}
This retrospective records a 100-line trilinear strict-improvement of
Tide 9. Tide 9 proved
$t^2 \cdot \kappa_3(x, x, x) \to -\alpha/\lambda^3$ for the 1D anharmonic
Gibbs measure with potential $L = (\lambda/2)x^2 + (\alpha/6)x^3 + (\gamma/24)x^4$.
This tide upgrades the result to arbitrary $\mathrm{id}$-multiples
$(b\,x, a\,x, c\,x)$, giving
$t^2 \cdot \kappa_3(b\,x, a\,x, c\,x) \to -(abc)\,\alpha/\lambda^3$,
via the static factorisation
$\kappa_3(b\,x, a\,x, c\,x) = (abc)\cdot\kappa_3(x, x, x)$.
The factorisation is hypothesis-free — it follows by trilinearity of
$\kappa_3$ once each of the seven Gibbs expectations in
$\mathtt{Threepoint.kappa3}$'s definition has its scalar factor peeled
off via I2's just-landed
\texttt{gibbsExpectation\_smul}.\footnote{I2 (gibbsCov-algebra) landed
2026-05-07. Branch: \texttt{tide/gibbscov-algebra}.}
The asymptotic is then $\mathtt{Tendsto.const\_mul}\,(abc)$ on top of
Tide 9. Zero \texttt{sorry}, single session, build green on the first
attempt; the GPT-5.5~Pro deliberation agreed with the Claude draft and
flagged a deferred shifted-affine extension.
\end{abstract}

\section{Setting}

Tide 9 (anharmonic-$\kappa_3$, 6 May) proved
\[
  t^2 \cdot \kappa_3(\mathrm{volume}, L_{\lambda\alpha\gamma}, x, t, x, x)
  \;\to\; -\alpha/\lambda^3
  \quad\text{as } t \to \infty
\]
for $L_{\lambda\alpha\gamma}(x) = (\lambda/2)x^2 + (\alpha/6)x^3 + (\gamma/24)x^4$,
under the discriminant condition $\alpha^2 < 3\lambda\gamma$.\footnote{Discriminant
condition guarantees the sextic Wick contractions in $\langle x^4\rangle_t$ stay
positive, so the asymptotic of the covariance $\langle x^2, x\rangle$ has the
sign Tide 9 records.} The slot tuple $(\mathrm{id}, \mathrm{id}, \mathrm{id})$
specialises $\kappa_3$'s general $(A, \varphi, B)$ signature to all-identity
observables. Tide 9 was the first cross-repo bridge between the abstract
\texttt{kappa3} machinery in \texttt{threepoint} and the concrete anharmonic
asymptotics in \texttt{laplace}; its retrospective named the affine-observable
form as the natural multilinear strict-improvement and estimated $\sim$40
lines.

I2 (gibbsCov-algebra, 7 May) landed an unconditional Gibbs-expectation /
Gibbs-covariance algebra:\footnote{Eleven lemmas mirrored across
\texttt{Laplace/Gibbs.lean} and \texttt{Laplace/Multi/Basic.lean}:
\texttt{gibbsExpectation\_\{smul,add,zero,const\}} and
\texttt{gibbsCov\_\{symm,smul,const,add,zero\}}.} Of these, the only
load-bearing one for this tide is
$\texttt{gibbsExpectation\_smul} : \langle c\,\varphi\rangle_t = c \cdot \langle \varphi\rangle_t$,
which holds unconditionally (when the partition is zero, both sides are
zero). The trilinearity factorisation reduces to seven applications of
this identity plus a \texttt{ring} step to collect the scalar factor
$abc$; no \texttt{gibbsCov} algebra is needed.

This tide was picked at 2026-05-07T20:55Z from candidate G2 in the v3
candidates survey,\footnote{\texttt{projects/automation/log/2026-05-07-v3-tide-candidates-survey.md}.}
after a parallel session claimed C1 (the harmonic-side twin) at 20:54Z.
Together with the in-progress C1, the two land complete affine-observable
$\kappa_3$ instances across both harmonic and anharmonic sides on the
same day.

\section{The thing formalised}

Two new public theorems in \texttt{Laplace/OneD/AnharmonicKappa3.lean}.

\paragraph{Theorem 1 (potential-agnostic factorisation).} For any
$L : \mathbb{R} \to \mathbb{R}$, $t \in \mathbb{R}$, and scalars
$a, b, c \in \mathbb{R}$,
\[
  \kappa_3(\mathrm{volume}, L,\; b \cdot x,\; t,\; a \cdot x,\; c \cdot x)
  \;=\;
  (a \cdot b \cdot c) \cdot \kappa_3(\mathrm{volume}, L,\; x,\; t,\; x,\; x).
\]
Lean signature:
\begin{quote}\small\ttfamily
theorem kappa3\_affine\_id\_id\_id\_eq\\
\hspace*{1em}(L : ℝ → ℝ) (t : ℝ) (a b c : ℝ) :\\
\hspace*{1em}Threepoint.kappa3 (volume) L\\
\hspace*{3em}(fun x => b * x) t (fun x => a * x) (fun x => c * x)\\
\hspace*{1em}= (a * b * c) * Threepoint.kappa3 (volume) L\\
\hspace*{6em}(fun x => x) t (fun x => x) (fun x => x)
\end{quote}
The factorisation has no analytic preconditions: $\kappa_3$'s seven
ratio-of-integrals reduce, at $h = 0$, to seven
\texttt{Laplace.gibbsExpectation} ratios via the existing
\texttt{threepoint\_gibbsExp\_volume\_zero\_eq} bridge, and
\texttt{gibbsExpectation\_smul} is unconditional.

\paragraph{Theorem 2 (anharmonic asymptotic).} For
$L_{\lambda\alpha\gamma}$ with $0 < \lambda$, $0 < \gamma$,
$\alpha^2 < 3\lambda\gamma$, and any $a, b, c \in \mathbb{R}$,
\[
  t^2 \cdot \kappa_3(\mathrm{volume}, L_{\lambda\alpha\gamma},\;
    b \cdot x,\; t,\; a \cdot x,\; c \cdot x)
  \;\to\; -(a \cdot b \cdot c) \cdot \alpha/\lambda^3
  \quad \text{as } t \to \infty.
\]
Lean signature:
\begin{quote}\small\ttfamily
theorem kappa3\_anharmonic\_affine\_asymptotic\\
\hspace*{1em}\{lam alpha gamma : ℝ\}\\
\hspace*{1em}(hlam : 0 < lam) (hgamma : 0 < gamma) (hdisc : alpha\^{}2 < 3 * lam * gamma)\\
\hspace*{1em}(a b c : ℝ) :\\
\hspace*{1em}Filter.Tendsto\\
\hspace*{2em}(fun t : ℝ => t\^{}2 * Threepoint.kappa3 (volume)\\
\hspace*{4em}(anharmonicPotential lam alpha gamma)\\
\hspace*{4em}(fun x => b * x) t (fun x => a * x) (fun x => c * x))\\
\hspace*{2em}Filter.atTop\\
\hspace*{2em}(nhds (-(a * b * c) * alpha / lam\^{}3))
\end{quote}
The hypotheses are exactly those of Tide 9: trilinearity of $\kappa_3$ in
the three observable slots costs nothing analytically.

\section{Proof strategy}

\paragraph{Theorem 1.} The key observation is that
$\mathtt{Threepoint.kappa3}$ unfolds to seven $\mathtt{gibbsExp}$
terms,\footnote{Specifically: $\langle\varphi A B\rangle$,
$\langle\varphi A\rangle\langle B\rangle$,
$\langle\varphi B\rangle\langle A\rangle$,
$\langle A B\rangle\langle\varphi\rangle$, and
$\langle\varphi\rangle\langle A\rangle\langle B\rangle$ — a polynomial
of total degree at most three in the seven scalar values.} all
evaluated at $h = 0$. The bridging lemma
\texttt{threepoint\_gibbsExp\_volume\_zero\_eq} (already in laplace,
from Tide 9) routes each one through to
\texttt{Laplace.gibbsExpectation}, after which the seven integrand
$\lambda$-binders carry products like
$(\mathtt{fun}\;x \mapsto a\,x)\,w \cdot (\mathtt{fun}\;x \mapsto b\,x)\,w$
that are not yet syntactically in the canonical scaled-monomial form.
A \texttt{funext;\,ring} cleanup\footnote{Seven such
\texttt{have h\_X : (fun w => ...) = (fun w => K * w\^{}n)} steps:
$abc \cdot w^3$, $ab \cdot w^2$, $ac \cdot w^2$, $bc \cdot w^2$, plus
the three $a w$, $b w$, $c w$ singletons, and the four RHS-side
identity-only forms $w^3$ and $w^2$.} forces them into the canonical
shapes. \texttt{gibbsExpectation\_smul} then peels each scalar factor
out of its Gibbs expectation, leaving a pure ring identity in
$(a, b, c) \in \mathbb{R}^3$ and the four scalar values
$\langle w^3 \rangle$, $\langle w^2 \rangle$, $\langle w \rangle$, $\langle w \rangle^3$
(which appear cubically in the standard polarisation
$\langle w^3 \rangle - 3\langle w^2\rangle\langle w\rangle + 2\langle w\rangle^3$).
\texttt{ring} closes.

\paragraph{Theorem 2.} The new \texttt{Tendsto} is just $abc$ times
Tide 9's: $t^2 \cdot \kappa_3(b\,x, a\,x, c\,x) = abc \cdot t^2 \cdot \kappa_3(x, x, x)$
pointwise. So the proof is
$\mathtt{Tendsto.const\_mul}\,(abc)$ applied to the Tide 9 conclusion,
combined with $\mathtt{Tendsto.congr'}$ rewriting the function under
the integral via Theorem 1. The constant
$-(a b c)\,\alpha/\lambda^3$ is normalised against
$abc \cdot (-\alpha/\lambda^3)$ by a one-line \texttt{ring}.

\section{Roadblocks and resolutions}

There were no roadblocks. The proof landed green on the first
\texttt{lake build}, in 4.7\,s after a warm Mathlib cache. This is
unusual but reasonable for a tide whose math is pure trilinearity over
a closed-form cumulant; the load-bearing analytic content is fully
imported from Tide 9.

The deliberation step ran in parallel with the formalisation in this
session, on the assumption (correct, as it turned out) that the
candidate was uncontroversial. GPT-5.5~Pro's response, when it
returned 4 minutes after the proof had already landed, confirmed
Candidate~A as the right patch-sized theorem and noted three small
points worth recording:

\begin{enumerate}
\item \emph{The asymptotic constant.} GPT confirmed the sign
  $-(abc)\,\alpha/\lambda^3$ (no factor or sign error). Suggested
  stating the limit as \texttt{((a * b * c) * (-α / λ\^{}3))} for
  Lean ergonomics; the proof uses
  \texttt{(-(a * b * c) * alpha / lam\^{}3)} and bridges the two
  forms by a single \texttt{ring} step. Both work; one wonders if
  Mathlib will eventually grow a normal-form simp set for these.
\item \emph{Algebra path.} ``You likely don't need \texttt{gibbsCov}
  for this patch; direct \texttt{gibbsExpectation\_smul} +
  \texttt{gibbsExpectation\_const} is enough.'' Exactly the route
  taken; the I2 \texttt{gibbsCov} algebra is not used here. (It is
  used in C1, the harmonic-side twin.)
\item \emph{Missed candidate (deferred).} ``The affine
  strengthening is also true:
  $\kappa_3(b\,x + d, a\,x + e, c\,x + f) = abc \cdot \kappa_3(x, x, x)$,
  because \texttt{kappa3} kills constants in each slot
  ($\mathbb{E}[1] = 1$ makes the cancellations exact). But I would
  defer it unless you need shifted observables now.'' Recorded as a
  follow-up; not picked up here.
\end{enumerate}

\section{What was Mathlib, what was new}

\paragraph{Mathlib.} \texttt{Tendsto.const\_mul} (constant multiple of
a limit), \texttt{Tendsto.congr'} (rewriting the function under a
limit), and \texttt{filter\_upwards} (eventual quantification). All
appear in dozens of laplace files; nothing new from Mathlib.

\paragraph{Laplace seabed (existing).} Three load-bearing
imports:\footnote{Tide 9's
\texttt{kappa3\_anharmonic\_id\_id\_id\_asymptotic} (the headline),
its private bridge \texttt{threepoint\_gibbsExp\_volume\_zero\_eq},
and I2's \texttt{Laplace.gibbsExpectation\_smul}.}
Tide 9's headline asymptotic, Tide 9's $h = 0$ bridge to
\texttt{Laplace.gibbsExpectation}, and I2's unconditional
scalar-out-of-Gibbs-expectation lemma.

\paragraph{New.} Two public theorems and seven
\texttt{funext;\,ring}-style integrand-cleanup \texttt{have}-blocks
inside the proof of Theorem 1. Nothing new at the lemma library level.
The seven \texttt{funext}s are mechanical bookkeeping; a future tide
could fold them into a small simp set
``\texttt{push\_smul\_under\_lambda}'' but at one use it's not yet
cost-positive.

\section{Lessons}

\paragraph{Trilinearity tides land in one shot when the algebra is
already there.} I2 paid for itself the moment a strict-improvement of
this shape was on the board. The factorisation theorem is essentially
the I2 \texttt{gibbsExpectation\_smul} lemma applied seven times under a
\texttt{ring}-closure; if I2 hadn't been merged we'd have unfolded each
expectation manually. The lesson generalises to C1 (harmonic-side
twin) and is implicit in the I2 retrospective's follow-up list, which
flagged C1 and G2 specifically.

\paragraph{Parallel deliberation + formalisation is fine when the
deliberation outcome is predictable.} Auto mode here meant launching
the GPT consult in the background and writing the Lean code in the
foreground, without waiting. The risk is wasted work if GPT had named
a different candidate; the actual outcome (GPT's vote agreed with
Claude's) suggests this risk is acceptable for trilinearity-shaped
tides where the candidate is essentially forced. Less acceptable for
hypothesis-class deliberations or when the proof technique is
genuinely contested. Don't generalise this to all tides.

\paragraph{The ``\texttt{funext;\,ring} cleanup'' pattern is the
hidden cost of the seven-term \texttt{kappa3} unfolding.} Both Tide 9's
\texttt{kappa3\_id\_id\_id\_unfold} and this tide's
\texttt{kappa3\_affine\_id\_id\_id\_eq} carry the same boilerplate:
the seven Gibbs-expectation integrands need their lambdas reduced to
canonical scaled-monomial form before \texttt{gibbsExpectation\_smul}
can fire on them. A second use isn't yet a refactor trigger, but if a
third \texttt{kappa3} unfolding tide lands (e.g.\ for $\langle x^k \rangle$
slots, or for the cross-cumulant $\kappa_3(\varphi, A, B)$ of the
patterning paper), the pattern is well-developed enough to extract.

\section{Follow-ups}

\begin{itemize}
\item \emph{Shifted-affine strengthening} (GPT's missed candidate).
  $\kappa_3(b\,x + d, a\,x + e, c\,x + f) = abc \cdot \kappa_3(x, x, x)$
  via \texttt{gibbsExpectation\_const}-style cancellations
  ($\mathbb{E}[1] = 1$). $\sim$50 extra lines on top of this tide.
  Defer until a second consumer for shifted observables materialises.

\item \emph{C1 (harmonic-side twin), in flight.} The harmonic case
  $\kappa_3(b\,x, a\,x, c\,x) = abc \cdot \kappa_3(x, x, x) = 0$
  follows by the same trilinearity argument, on top of Tide 5's
  harmonic-vanishing theorem.\footnote{\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero}.}
  Claimed by a parallel session at 20:54Z; should land within hours
  of this retrospective. Together the two tides close the
  affine-observable $\kappa_3$ pair for the 1D potentials currently
  in laplace.

\item \emph{Generic multilinear factorisation (Candidate~C, deferred).}
  $\kappa_3(\mu, L, a A_0, t, b \varphi_0, c B_0) = abc \cdot \kappa_3(\mu, L, A_0, t, \varphi_0, B_0)$
  for arbitrary $A_0, \varphi_0, B_0$. Subsumes Theorem~1 as the
  $(\mathrm{id}, \mathrm{id}, \mathrm{id})$ instance. GPT noted ``only
  worth exporting if it is essentially free after the $h=0$ rewrite'';
  worth checking against C1's proof to see whether a unified C-shaped
  helper would shorten both files. If yes, a small refactor tide.

\item \emph{Promote ``\texttt{push\_smul\_under\_lambda} simp set''
  on third use.} The seven-\texttt{funext} pattern (this tide and
  Tide 9) is a candidate for automation but not yet cost-positive.
  Third use triggers.

\item \emph{\texttt{\textbackslash leanref}-tag the threepoint paper, pass 3.}
  This tide is a corollary of the threepoint paper's multilinearity-of-$\kappa_3$
  remark.\footnote{Roughly the slot-by-slot linearity of cumulants
  used implicitly in the patterning subtractions.} Once the next
  process-tide tagging pass is run (currently candidate K2), tag the
  multilinearity claim with \texttt{\textbackslash leanref} pinned to
  this commit.
\end{itemize}

\end{document}
