\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ŌäØ}{\ensuremath{\mathbb{R}}}
\newunicodechar{Ō¤©}{\ensuremath{\langle}}
\newunicodechar{Ō¤®}{\ensuremath{\rangle}}
\newunicodechar{╬ō}{\ensuremath{\Gamma}}
\newunicodechar{┬Ę}{\ensuremath{\cdot}}
\newunicodechar{┬▓}{\ensuremath{^2}}
\newunicodechar{┬│}{\ensuremath{^3}}
\newunicodechar{Ōü┤}{\ensuremath{^4}}
\newunicodechar{Žł}{\ensuremath{\psi}}
\newunicodechar{Žå}{\ensuremath{\varphi}}
\newunicodechar{╬▒}{\ensuremath{\alpha}}
\newunicodechar{╬╗}{\ensuremath{\lambda}}
\newunicodechar{Ōēż}{\ensuremath{\le}}

\setlength{\emergencystretch}{3em}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{Bounded-prior quartic partition function (grammar precursor)}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{2 May 2026}

\begin{document}
\maketitle

\begin{abstract}
A single Tide-loop excursion in the \texttt{laplace} seabed, formalising
the smallest possible step toward the grammar paper's
$Z_n[\varphi] = \int \varphi(w) e^{-n K(w)} \pi(w)\,dw$ machinery: replace
the integration domain $\mathbb{R}$ in the existing partition function
by a bounded interval $[-a,a]$ and quantify the resulting boundary
correction.\footnote{Headline theorem:
\texttt{Laplace.OneD.quartic\_partition\_bounded\_prior}.} The proof
adds 232 lines of Lean to the seabed. Zero \texttt{sorry}, zero
\texttt{axiom}, zero \texttt{native\_decide}; clean diagnostics on first
build after a parser fix on the absolute-value statement. The interesting
moves were upstream: a deliberation step that took an over-ambitious
direction (the full 2D $K = w_1^2 w_2^4$ resolution example, RLCT $3/4$
with logarithm) and pushed it down to a 1D one-ingredient-at-a-time
precursor, and a numerical sanity check that surfaced a loose exponent
constant in the candidate statement before any Lean was written.
\end{abstract}

\section{Setting}

This Tide step sits in the same \texttt{laplace} seabed used by five
preceding tide steps (the quartic moments, 2D semi-degenerate, universal
scaling, $f_1$ large-$t$, and 2D pure-quartic
excursions),\footnote{Branches: \texttt{tide/quartic-moments},
\texttt{tide/2d-semi-degenerate}, \texttt{tide/universal-scaling},
\texttt{tide/f1-large-t}, \texttt{tide/2d-pure-quartic}.} but it is the
first tide aimed not at the Susceptibility Primer's Section~4 but at the
\emph{grammar paper} (Gerraty et al., \emph{Expectations and the
Exceptional Divisor}). Per the cross-tide candidates
survey,\footnote{\texttt{projects/automation/log/2026-05-02-tide-candidates-survey.md}}
the grammar paper's \emph{simplest resolution example}\,---\,$K(w) =
w_1^2 \cdot w_2^4$ on a 2D parameter space with bounded prior, leading
asymptotic $Z_n[\varphi] \sim C_\varphi \cdot n^{-3/4} \log n$ with RLCT
$\lambda = 3/4$ and multiplicity $m = 2$\,---\,is the natural next
target, but the survey itself flagged the full 2D form as ``probably the
wrong shape for a single excursion'' and asked Step~2 to pick a 1D
precursor instead.

The Step~2 deliberation drafted three candidates of varying ambition:
(A) bounded-prior quartic partition function with explicit exponential
remainder; (B) the same with a harmonic potential (essentially
glue-of-existing-seabed-lemmas); (C) bounded-prior quartic with a
continuous test function (the genuine 1D Watanabe asymptotic at a
regular critical point). GPT-5.5 Pro and Claude both backed A, on
matched grounds: B is too small to count as a step outward (the harmonic
tail is already in the seabed and there is no new quartic-side analysis
to do), C is too large because it bundles bounded-prior, test-function
localisation, change-of-variables, and dominated convergence in one
step. A introduces \emph{exactly one} ingredient (bounded prior, with
the necessary new quartic tail bound) on the most-recently-finished
tide step's potential. The full deliberation, with the GPT consult
preserved verbatim alongside, lives in the grammar project's
tide-log.\footnote{\texttt{projects/grammar/tide-log/2026-05-02-tide-grammar-precursor.md}
and \texttt{gpt55\_tide\_grammar\_v1.md} alongside it.}

The branch is chained on top of the prior 2D pure-quartic tide's
tip\footnote{Branch \texttt{tide/grammar-precursor-bounded-prior} on top
of \texttt{tide/2d-pure-quartic} at commit \texttt{efafbfa}.} per the
linear-chain branch-hygiene rule, so a single SHA-pin covers all
in-flight tides on the way to a squash-merge.

\section{The thing formalised}

For the centred quartic potential
\[
  L(w) = \frac{w^4}{24}
\]
and any $a > 0$, $t > 0$, the bounded-prior partition function differs
from the existing full-line \texttt{quartic\_partition} by an
exponentially small correction:

\begin{quote}\itshape
\[
  \left| \int_{-a}^{a} e^{-t \cdot w^4/24}\, dw \;-\;
        \tfrac{1}{2}\,(24/t)^{1/4}\,\Gamma(1/4) \right|
  \;\le\; \frac{24}{t \cdot a^3}\,
          \exp\!\left(-\tfrac{t \cdot a^4}{24}\right).
\]
\end{quote}

\noindent The Lean signature:
\begin{verbatim}
theorem quartic_partition_bounded_prior {t a : ŌäØ} (ht : 0 < t) (ha : 0 < a) :
    |(Ōł½ w in Icc (-a) a, Real.exp (-(t * w ^ 4 / 24))) -
        (1/2) * (24/t) ^ ((1 : ŌäØ) / 4) * Real.Gamma ((1 : ŌäØ) / 4)| Ōēż
      (24 / (t * a ^ 3)) * Real.exp (-(t * a ^ 4 / 24))
\end{verbatim}

\noindent located at \texttt{Laplace/OneD/QuarticBoundedPrior.lean}, in
namespace \texttt{Laplace.OneD}.

This is the simplest possible step toward the grammar paper's
$Z_n[\varphi]$ machinery. The grammar paper's partition function differs
from the existing seabed in three ways: (i) bounded prior $\pi$ replacing
the full real line, (ii) test function $\varphi$ inside the integrand,
(iii) explicit asymptotic in $n$ with remainder. The present theorem
adds \emph{only} (i), on the most degenerate 1D potential we already
have, with a remainder that is not just asymptotic but \emph{exact} for
all $t > 0$. Subsequent tide steps can layer (ii) and the monomial
generalisation on top, with the bounded-prior tail-bound infrastructure
already in place.

Numerical sanity check: \texttt{scipy.integrate.quad} on $[-a,a]$ for
$(a,t) \in \{1,2\} \times \{10, 100, 1000\}$ agreed with leading
$+$ tail bound to machine precision (e.g.\ at $a=1, t=100$:
numeric $1.2672$, leading $1.2688$, error $1.6\times 10^{-3}$, tail
bound $3.7\times 10^{-3}$). The same script also validated candidate B
(harmonic) and candidate C (cosine test function, ratio $\to 1$ slowly
as $t\to\infty$); recorded in the tide log.

\section{Proof strategy}

The proof has three pieces: a tail sublemma, a tail-symmetry lemma, and a
splitting identity that reduces the headline to combining them.

The \emph{key sublemma} is \texttt{quartic\_tail\_Ioi}:
\[
  \int_a^\infty e^{-t \cdot w^4/24}\, dw
  \;\le\; \frac{12}{t \cdot a^3}\,\exp\!\left(-\tfrac{t \cdot a^4}{24}\right).
\]
The proof reduces the quartic tail to the existing Gaussian tail. For
$w \ge a$ we have $w^2 \ge a^2$, hence $w^4 = w^2 \cdot w^2 \ge a^2 \cdot
w^2$, hence
\[
  e^{-(t/24) w^4} \le e^{-(t a^2/24) w^2}
  = e^{-(b/2) w^2} \quad\text{with } b = t a^2/12.
\]
Apply the seabed's \texttt{Laplace.OneD.gaussian\_tail\_bound\_rescaled\_Ioi}
at $M = a$, $b = t a^2/12$:
\[
  \int_a^\infty e^{-(b w^2)/2}\, dw \le \frac{e^{-b a^2 / 2}}{b \cdot a}
  = \frac{12}{t a^3}\,\exp\!\big(-\tfrac{t a^4}{24}\big).
\]
That bound is already what we want. The Lean form glues a set-integral
monotonicity lemma on the pointwise bound to the seabed's Gaussian tail
bound,\footnote{\texttt{setIntegral\_mono\_on}, applied to
\texttt{gaussian\_tail\_bound\_rescaled\_Ioi}.} with integrability of
the quartic side from the seabed's \texttt{quartic\_integrable\_pow}
and of the Gaussian comparator from Mathlib's
\texttt{integrable\_exp\_neg\_mul\_sq}.

The \emph{tail-symmetry lemma} mirrors the seabed's Gaussian-side
analogue, using Mathlib's negation-substitution lemma together with the
evenness identity $(-w)^4 = w^4$.

The \emph{splitting identity} writes $\mathbb{R}$ as
$\text{Iic}\,a \cup \text{Ioi}\,a$ via Mathlib's
\texttt{integral\_Iic\_add\_Ioi}, then $\text{Iic}\,a$ as
$\text{Iio}\,(-a) \cup \text{Icc}\,(-a,a)$ via
\texttt{setIntegral\_union} with a hand-written disjointness witness
and a six-line \texttt{ext}/\texttt{rcases} union proof.\footnote{The
new module exports this as \texttt{quartic\_partition\_split}.}
Combined with the tail symmetry, this gives
\[
  \int_\mathbb{R} e^{-t L} \;=\; \int_{[-a,a]} e^{-t L}
                                \;+\; 2\int_{(a,\infty)} e^{-t L}.
\]

The \emph{headline} then unfolds in eight lines: the full-line integral
equals \texttt{quartic\_partition} (via
\texttt{quartic\_moment\_even} at $n = 0$, after a small
\texttt{Nat.cast\_zero}\,/\,\texttt{mul\_zero}\,/\,\texttt{zero\_add} cleanup);
the absolute-value bound on the difference equals $2 \cdot
\int_{(a,\infty)}$, which by \texttt{quartic\_tail\_Ioi} doubled is
$(24/(t a^3)) \cdot \exp(-t a^4/24)$.

The total weight of the proof is concentrated in the
\texttt{setIntegral\_mono\_on} step (where the integrability bookkeeping
costs more than the mathematical bound itself) and the union-and-disjointness
glue for \texttt{setIntegral\_union}. Both are mechanical once the
strategy is right.

\section{Roadblocks and resolutions}

The deliberation surfaced two issues before any Lean was written, and
the formalisation surfaced two more.

\paragraph{Loose exponent in the original draft.}
Claude's first version of candidate~A had the remainder bound
$|r_a(t)| \le C_a \cdot \exp(-t a^4/48)$, derived from a
square-completion comparison $w^4/24 - w^2 \ge -6/t$. The factor $48$ is
correct but loose by a factor of $2$ in the exponent and folds the $t$
into an unstructured constant $C_a$. The numerical sanity check
immediately showed the bound was $\sim 100\times$ slack at $a=1, t=100$,
and GPT-5.5 Pro independently flagged the cleaner alternative:

\gptbox{A \emph{cleaner and sharper} bound is available:
$|r_a(t)| = 2 \int_a^\infty e^{-t w^4 / 24}\,dw \le \frac{24}{t a^3} e^{-t a^4 / 24}$.
This comes from the simple comparison $w^4 \ge a^2 w^2$ for $w \ge a$,
reducing the quartic tail to the already-formalised Gaussian tail.}

That same paragraph turned into the proof strategy. Calibration: the
numerical check sanity-checked both the leading constant and the
remainder shape, and the deliberation step caught the tightenable
exponent before any Lean compilation. Two reviewer cycles saved.

\paragraph{Disagreement on candidate scope.}
Both parties backed A, but GPT also flagged a Lipschitz-near-zero
middle-ground candidate (essentially candidate~C with weakened
hypothesis on $\varphi$) as worth considering. We deferred: the rule
``one ingredient at a time'' wins over ``cleanest single statement that
combines two ingredients.'' A future tide can pick up the test-function
side from this base.

\paragraph{Lean parser strict on \texttt{|} placement.}
The first version of the headline statement was written
\verb!|((Ōł½ ...) - leading) | Ōēż ...! with a space before the closing
\verb!|!. Lean's parser failed at column~67 with ``expected no space
before''. The fix was to drop the redundant outer parens:
\verb!|(Ōł½ ...) - leading| Ōēż ...!. The lesson is that the absolute-value
notation is more sensitive to surrounding whitespace than other
delimiter-like operators in Lean; in doubt, keep the inside terse.

\paragraph{Concurrent-agent staging race (recurrence).}
The repository hosting this Tide step had two other Tide branches
running concurrently (the 2D pure-quartic and sextic-moments
excursions).\footnote{Branches: \texttt{tide/2d-pure-quartic} and
\texttt{tide/sextic-moments}.} The top-level import file
\texttt{Laplace.lean} was rewritten by the parallel agents twice during
the run, each time stripping or reordering imports to match what was on
\emph{their} branch's \texttt{HEAD}. The fix in this run was to
(a) stash the working tree, (b) check out the import file from
\texttt{main} to a known baseline, (c) re-apply only the imports
relevant to this branch and the prior chained-base branch, then
(d) force-retarget the branch label onto the prior tide's tip before
committing.

This is the same staging race called out in the 2D pure-quartic
retrospective. The structural fix is independent of either tide:
concurrent agents need filesystem isolation via \texttt{git worktree}.
That's repeated under \emph{Lessons} below.

Beyond these four, no proof-side roadblocks. All four sub-theorems
typechecked clean within ten minutes once the strategy was pinned.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
The Gaussian integrability lemma for the comparator, the set-integral
monotonicity lemma for pulling a pointwise inequality through, the two
splitting lemmas (\texttt{Iic\,$\cup$\,Ioi} and the
\texttt{setIntegral\_union} disjoint-set decomposition), the negation
substitution \texttt{integral\_comp\_neg\_Iic} for the tail symmetry,
plus the standard \texttt{Real.exp\_le\_exp},
\texttt{mul\_self\_le\_mul\_self}, \texttt{pow\_pos}, and the
\texttt{ring}\,/\,\texttt{field\_simp}\,/\,\texttt{linarith}
closers.\footnote{Specifically: \texttt{integrable\_exp\_neg\_mul\_sq},
\texttt{setIntegral\_mono\_on}, and
\texttt{intervalIntegral.integral\_Iic\_add\_Ioi}.}

\paragraph{The seabed provided.}
The leading constant \emph{is} the existing 1D quartic partition
function lemma\,---\,no new closed-form computation
needed.\footnote{\texttt{Laplace.OneD.quartic\_partition} from
\texttt{Quartic.lean}.} The Gaussian tail in
\texttt{TailBound.lean}\footnote{\texttt{gaussian\_tail\_bound\_rescaled\_Ioi},
originally written for the harmonic case and used in
\texttt{Localisation.lean}.} is the input to which the quartic tail
reduces; it is reused here unchanged. Integrability of polynomials
against the quartic Boltzmann factor came from the existing
\texttt{quartic\_integrable\_pow} (also unchanged).

\paragraph{What was new in this tide.}
Four theorems in the new \texttt{QuarticBoundedPrior.lean} module:
a quartic half-line tail bound, a tail-symmetry lemma, a
partition-function splitting identity, and the headline bounded-prior
bound. None of the four theorems
introduces a new mathematical technique; they wire existing seabed and
Mathlib pieces in a new arrangement. The honest line-count breakdown is
roughly: 5 lines of new mathematics (the comparison
$w^4 \ge a^2 w^2$ and its consequence), 30 lines of arithmetic
manipulation (rewriting $b = t a^2/12$, computing
$\frac{1}{b \cdot a} = 12/(t a^3)$, etc.), and 200 lines of integrability
and measurability bookkeeping plus splitting/union proofs. That ratio
is the dominant cost on a small bounded-prior tide step in this style.

\section{Lessons}

\begin{itemize}
\item \emph{Numerical sanity check pays for itself before writing Lean.}
The exponent-constant tightening in the present run came purely from the
combination of \texttt{scipy.integrate.quad} numbers and a five-minute
GPT consult; it cost nothing to find and cost a factor-of-two reduction
in the final bound. Worth doing on every tide whose closed form is
numerical-friendly. The discipline already lives in the Tide skill;
record this run as another data point that confirms it.

\item \emph{Concurrent Tide agents need filesystem isolation.}
\texttt{Laplace.lean} is again the contention point; the situation is
identical to the one called out in the 2D pure-quartic retrospective.
Until \texttt{git worktree} support lands in the Tide skill's preflight,
expect to spend a few minutes per concurrent run on the recovery dance:
checkout the import-list file from \texttt{main}, re-add only the
imports relevant to this tide, and force-retarget the branch label onto
the prior tide's tip before committing. Documenting the dance here
should at least make the pattern recognisable on first hit in a future
run.

\item \emph{The substitution-and-Gaussian-tail template is reusable.}
The exact same proof pattern (pointwise comparison
$w^{2k} \ge a^{2k-2} \cdot w^2$, then the seabed's rescaled-Gaussian
tail bound) will give a one-line generalisation of the quartic tail
lemma to a monomial-of-arbitrary-even-degree tail lemma for
$L = w^{2k}/(2k)!$. GPT-5.5 Pro flagged this in the deliberation step
(``If you generalise, generalise the tail lemma first, not the full
bounded-prior asymptotic theorem''); worth acting on the next time the
grammar-precursor tide is taken further.

\item \emph{Lean parser is whitespace-sensitive on \texttt{|┬Ę|}.}
For the seabed's \texttt{CLAUDE.md}: when writing absolute-value
statements, prefer \verb!|expr| Ōēż ...! with no extra outer parens
between \verb!expr! and the closing \verb!|!. Lean's tokeniser treats
\verb!) |! as ambiguous in some positions.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{Bounded-prior quartic with continuous test function}
(candidate~C from this deliberation). Now the next-step grammar
precursor: bounded prior $+$ tail bound infrastructure is in place; what
remains is the localisation argument for $\varphi$ near zero plus the
\texttt{Asymptotics.IsLittleO} packaging. Estimated 400--600 lines.

\item \emph{Bounded-prior monomial generalisation}
$\big| \int_{-a}^{a} e^{-t w^{2k}/(2k)!}\,dw - \tfrac{1}{2}((2k)!/t)^{1/(2k)}
\Gamma(1/(2k)) \big| \le C_k(a)\,\exp(-c_k\,t a^{2k})$. The tail-bound
proof transfers verbatim, using $w^{2k} \ge a^{2k-2}\,w^2$ for $w \ge a$.
The hard part is the \texttt{Real.rpow} bookkeeping that comes with the
generalised exponent.

\item \emph{Bounded-prior harmonic partition} (candidate~B). Now
trivial given the seabed's harmonic partition lemma and the splitting
idiom from this tide. $\sim$\,50--100 lines if anyone wants the literal
Gaussian version on the record.

\item \emph{Worktree-isolated Tide runs.} Repeated from the 2D
pure-quartic retrospective. The recurrence here makes the case
stronger.

\item \emph{Promote the splitting/union idiom to \texttt{lean-common}.}
The hand-written \texttt{Iio\,(-a) \(\cup\) Icc\,(-a,a) = Iic\,a}
disjointness $+$ union proof is a self-contained, generic statement that
will recur on every bounded-prior tide. Six lines is small enough to
duplicate once but it should not duplicate three times.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step~2 and
the strategic move that became the final proof strategy: the
quartic-tail sublemma via the comparison $w^4 \ge a^2 w^2$. The full
consult and the tide log itself (with candidate drafts, vote, numerical
check, and result) are preserved alongside one another in the grammar
project's tide-log directory under the SRI repo.

\end{document}
