\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{ℓ}{\ensuremath{\ell}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{²}{\ensuremath{^2}}
\newunicodechar{³}{\ensuremath{^3}}
\newunicodechar{⁴}{\ensuremath{^4}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{π}{\ensuremath{\pi}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{∫}{\ensuremath{\int}}
\newunicodechar{∂}{\ensuremath{\partial}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{₀}{\ensuremath{_0}}
\newunicodechar{𝓝}{\ensuremath{\mathcal{N}}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{full-line J-function asymptotic for the quartic potential}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{7 May 2026}

\begin{document}

\sloppy
\maketitle

\begin{abstract}
A single-session Tide-loop excursion in the \texttt{laplace} seabed,
formalising Step 2 of the grammar paper's precursor sequence: the
full-line J-function asymptotic for the single-component blow-up
$K(w) = w^4/24$. Two theorems committed: a centred DCT theorem
$t^{1/4} \int e^{-t w^4/24}(\pi(w) - \pi(0))\,dw \to 0$ (Candidate X
from the deliberation), and the user-facing form $t^{1/4} J(t) \to
\pi(0) \cdot \tfrac12 \cdot 24^{1/4} \cdot \Gamma(1/4)$ as a 2-line
corollary using \texttt{quartic\_partition}. 280 lines of Lean over
the existing seabed, zero \texttt{sorry}, single-session
formalisation. The deliberation reshaped the target from the direct
form (Candidate~A) to the centred form (X)\,---\,less algebra in the
statement, the user-facing form falls out\,---\,which Claude initially
preferred but GPT correctly pushed harder for. The K1 worktree
discipline (landed earlier today) gave clean filesystem isolation;
no collisions with the four other concurrent Tide branches.
\end{abstract}

\section{Setting}

The grammar paper precursor sequence is a multi-tide programme aimed
at building enough infrastructure in the \texttt{laplace} seabed to
formalise the resolution-of-singularities argument that drives the
grammar paper's main theorem. The sequence's published steps so far:

\begin{itemize}[leftmargin=*]
\item \emph{Step 1} (Tide 7): bounded-prior partition function for
the quartic\,---\,$Z_a(t) := \int_{[-a,a]} e^{-tw^4/24}\,dw$, with
positivity, tail bound, and the splitting infrastructure that
subsequent precursor steps reuse.
\item \emph{Step 2-prime} (Tide 11+): bounded-prior with continuous
test function. Branch-side closure has primitives plus a Lipschitz
unnormalised estimate; the main-branch merge has only the primitives.
\end{itemize}

This Tide is the next link in that chain: drop the bounded-prior
restriction and prove the \emph{full-line} J-function asymptotic.
Concretely, for the single-monomial potential $K(w) = w^4/24$ and a
continuous bounded prior $\pi : \mathbb{R} \to \mathbb{R}$, prove
\[
  t^{1/4} \cdot J(t) \;:=\; t^{1/4} \int_\mathbb{R} e^{-t\,w^4/24}\,\pi(w)\,dw
  \;\longrightarrow\; \pi(0) \cdot \tfrac12 \cdot 24^{1/4} \cdot \Gamma(1/4)
  \quad\text{as } t \to \infty.
\]

The Step-2 deliberation\footnote{See \texttt{projects/primer/tide-log/2026-05-07-tide-grammar-j-function.md}
in the SRI repo for the full deliberation transcript and Candidates v1.}
drafted three candidates spanning scope (k=2 only / generic k /
bounded-support variant) and converged on a fourth, GPT-proposed
reshape: the \emph{centred} form
$t^{1/4}\int e^{-tw^4/24}(\pi(w)-\pi(0))\,dw \to 0$, with the
user-facing direct form derived as a 2-line corollary using
\texttt{quartic\_partition} (Tide 7's closed form for the no-prior
partition).

\section{The thing formalised}

\begin{quote}\itshape
For $K(w) = w^4/24$ and $\pi : \mathbb R \to \mathbb R$ continuous with
$|\pi(w)| \le M$ for some $M$,
\[
  \lim_{t \to \infty} t^{1/4} \int_{\mathbb R} e^{-t\,w^4/24}\,(\pi(w) - \pi(0))\,dw \;=\; 0,
\]
and consequently
\[
  \lim_{t \to \infty} t^{1/4} \int_{\mathbb R} e^{-t\,w^4/24}\,\pi(w)\,dw
  \;=\; \pi(0) \cdot \tfrac12 \cdot 24^{1/4} \cdot \Gamma(1/4).
\]
\end{quote}

\noindent The Lean signatures, in namespace \texttt{Laplace.OneD}:
\begin{verbatim}
theorem quartic_jfunction_centered_tendsto_zero
    {prior : ℝ → ℝ} (hprior_cont : Continuous prior)
    {M : ℝ} (hprior_bd : ∀ x, |prior x| ≤ M) :
    Tendsto (fun t : ℝ =>
        t ^ ((1:ℝ)/4) *
          ∫ w : ℝ, Real.exp (-(t * w^4 / 24)) * (prior w - prior 0))
      atTop (𝓝 0)

theorem quartic_jfunction_asymptotic
    {prior : ℝ → ℝ} (hprior_cont : Continuous prior)
    {M : ℝ} (hprior_bd : ∀ x, |prior x| ≤ M) :
    Tendsto (fun t : ℝ =>
        t ^ ((1:ℝ)/4) *
          ∫ w : ℝ, Real.exp (-(t * w^4 / 24)) * prior w)
      atTop (𝓝 (prior 0 * (1/2) * (24:ℝ) ^ ((1:ℝ)/4) *
                Real.Gamma (1/4)))
\end{verbatim}

\noindent located at \texttt{Laplace/OneD/JFunctionQuartic.lean}.

The hypothesis `Continuous prior` is slightly stronger than the
deliberation's `AEMeasurable + ContinuousAt 0 + bounded`. The Lean
saving (~50 lines of \texttt{AEMeasurable.comp\_quasiMeasurePreserving}
plumbing) was worth it for a Tide step; the weakening is recorded as
a follow-up. All paper applications use continuous priors anyway.

\section{Proof strategy}

The proof is substitute-then-DCT. With $a := t^{1/4}$ (so $a^4 = t$),
the change of variables $u = a \cdot w$ converts
\[
  t^{1/4} \int_\mathbb{R} e^{-t\,w^4/24}\,(\pi(w) - \pi(0))\,dw
\]
into the $t$-independent-integrand form
\[
  \int_\mathbb{R} e^{-u^4/24}\,(\pi(u/a) - \pi(0))\,du.
\]

The change of variables is implemented by Mathlib's
\texttt{MeasureTheory.Measure.integral\_comp\_mul\_right}, which says
$\int g(w \cdot a)\,dw = |a^{-1}| \cdot \int g(u)\,du$. Setting
$g(u) := e^{-u^4/24}\,(\pi(u/a) - \pi(0))$, we have
$g(w \cdot a) = e^{-t\,w^4/24}\,(\pi(w) - \pi(0))$ (using
$a^4 = t$ and $(w \cdot a)/a = w$), so the original integral equals
$|a^{-1}| \cdot \int g(u)\,du = (1/a) \int g(u)\,du$. Multiplying by
$a = t^{1/4}$ gives the substituted form.

The DCT step uses
\texttt{MeasureTheory.tendsto\_integral\_filter\_of\_dominated\_convergence}
with envelope $2M\,e^{-u^4/24}$ (integrable by Tide 7's
\texttt{quartic\_integrable\_pow\,0}) and pointwise limit $0$
(from $u/a \to 0$ as $a \to \infty$ plus continuity of $\pi$ at the
origin, then multiplication by the bounded $e^{-u^4/24}$).

The user-facing corollary follows by splitting
$\pi(w) = (\pi(w) - \pi(0)) + \pi(0)$ inside the integral. The
centred piece is the headline; the constant piece is
$\pi(0) \cdot t^{1/4} \cdot \int e^{-t w^4/24}\,dw$, evaluated by
Tide 7's \texttt{quartic\_partition} closed form. Multiplying out
$t^{1/4} \cdot (24/t)^{1/4} = 24^{1/4}$ yields the constant
$\pi(0) \cdot \tfrac12 \cdot 24^{1/4} \cdot \Gamma(1/4)$.

The proof body is structured as: a private substitution lemma
\texttt{jfunction\_centered\_subst} (~55 lines, the most
arithmetically dense step\,---\,handles the
\texttt{Real.rpow} Jacobian bookkeeping) and a small dominator-
integrability helper, followed by the headline theorem (~50 lines,
DCT plus the substitution rewrite) and the corollary (~70 lines,
integral split + closed-form substitution).

\section{Roadblocks and resolutions}

\paragraph{`π` clashes with Lean 4.29's namespace.} The first
skeleton bound the prior as `\{π : ℝ → ℝ\}`. Lean rejected: \texttt{unexpected token 'π'; expected '\_' or
identifier}. The Greek letter `π` is somehow treated as an operator
or reserved-ish in this Lean version. Rename to \texttt{prior}; the
rebuild went through. Worth landing in \texttt{laplace/CLAUDE.md} as
a v4.29 quirk.

\paragraph{\texttt{Real.tendsto\_rpow\_atTop} is at root namespace.}
First attempt at the pointwise limit step used
\texttt{Real.tendsto\_rpow\_atTop}. Build error: \texttt{Unknown
constant `Real.tendsto\_rpow\_atTop`}. The lemma lives at the root
namespace as \texttt{tendsto\_rpow\_atTop} (in
\texttt{Mathlib.Analysis.SpecialFunctions.Pow.Asymptotics}). This is
the same gotcha pattern as
\texttt{MeasureTheory.Measure.integral\_comp\_mul\_right}\,---\,namespace
inconsistency between morally-related lemmas. Worth a one-liner in
\texttt{CLAUDE.md}.

\paragraph{`.symm` on `\texttt{∀ᶠ}` doesn't dot-resolve to `Filter.EventuallyEq.symm`.}
The combine-via-Tendsto.congr' step originally wrote
\texttt{Tendsto.congr' h\_decomp.symm h\_sum\_C}. Lean reported
\texttt{Invalid field notation: ... `Filter.Eventually` is not a
type that has a `symm` projection}. The fix is
\texttt{h\_decomp.mono fun \_ h => h.symm} (use
\texttt{Filter.Eventually.mono} to symmetrise the inner equality
pointwise) or explicit
\texttt{Filter.EventuallyEq.symm h\_decomp}. Subtle because in prose
\texttt{∀ᶠ t in atTop, f t = g t} \emph{is} \texttt{f =ᶠ[atTop] g}\,---\,but
the \texttt{def} unfolding doesn't propagate to \texttt{.symm} dot
resolution. Worth a CLAUDE.md note.

\paragraph{Hypothesis weakening: \texttt{AEMeasurable} → \texttt{Continuous}.}
The deliberation chose \texttt{AEMeasurable + ContinuousAt 0 + global bound}
as the mathematically-minimal hypothesis; the formalisation
strengthened to \texttt{Continuous}. Reason: the AEMeasurable case
needs \texttt{AEMeasurable.comp\_quasiMeasurePreserving} for the
composition $\pi \circ (\cdot/a)$, plus \texttt{AEStronglyMeasurable}
follow-through, plus a similar argument inside the DCT bound step\,---\,collectively
~50 lines of MeasureTheory plumbing. The strengthened hypothesis
captures all paper applications and saves a Tide-substantial chunk
of proof. The weakening is recorded as a follow-up tide; nothing in
the existing proof obstructs the substitution.

\paragraph{No proof-side roadblocks beyond the above.} The
substitution lemma typechecked first try once the \texttt{a^4 = t}
identity was set up correctly (the trick: convert to
\texttt{Real.rpow} via \texttt{Real.rpow\_natCast} then use
\texttt{Real.rpow\_mul}). The DCT and corollary went through with
modest tactic adjustments\,---\,no thrashing.

\paragraph{Worktree isolation worked as intended.} This Tide ran in
a dedicated \texttt{git worktree} created via
\texttt{tide-worktree create laplace grammar-j-function} (the K1
helper landed earlier today). Four other Tide branches
were active concurrently in their own worktrees; no
\texttt{Laplace.lean} import-line clobber, no stray-message commits,
no branch-toggle file disappearances. The worktree discipline
demonstrably solves the failure class the 6 May
cross-susc-deriv-harmonic retrospective (Tide 13) flagged as
``overdue.''

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
\texttt{MeasureTheory.tendsto\_integral\_filter\_of\_dominated\_convergence}
for the DCT step\footnote{The \texttt{atTop : Filter ℝ} is countably
generated, so the IsCountablyGenerated typeclass slot is filled
automatically.}.
\texttt{MeasureTheory.Measure.integral\_comp\_mul\_right} for the
change of variables (with the Jacobian factor $|a^{-1}|$).
\texttt{tendsto\_rpow\_atTop} (with the Real-namespace gotcha
above), \texttt{Tendsto.div\_atTop}, \texttt{Tendsto.const\_mul},
\texttt{Tendsto.sub\_const} for the pointwise-limit chain.
\texttt{Real.rpow\_natCast}, \texttt{Real.rpow\_mul},
\texttt{Real.mul\_rpow} for the
$t^{1/4}$ bookkeeping. \texttt{MeasureTheory.integral\_add},
\texttt{integral\_mul\_const} for the corollary's split.

\paragraph{Tide 7 (laplace seabed) provided.}
\texttt{quartic\_partition} (the closed form
$\int e^{-t\,w^4/24}\,dw = \tfrac12 (24/t)^{1/4} \Gamma(1/4)$) and
\texttt{quartic\_integrable\_pow} (envelope integrability for the
DCT bound) are the only seabed pieces used.

\paragraph{What was new.} The substitution lemma
\texttt{jfunction\_centered\_subst}\,---\,a $t^{1/4}$-prefixed change
of variables that turns the original integrand into a
$t$-independent shape. The dominator-integrability helper. The two
headline theorems. ~280 lines total.

\paragraph{Out of scope.} The generic-$k$ version (for
$K = w^{2k}/(2k)!$ at $k = 1, 2, 3, \ldots$). The hypothesis
weakening to \texttt{AEMeasurable + ContinuousAt 0}. The
asymptotic-equivalence form $J(t) \sim \pi(0) Z(t)$ (rather than the
\texttt{Tendsto} form), which is one
\texttt{Tendsto.div}-style step from what we have.

\section{Lessons}

\begin{itemize}[leftmargin=*]
\item \emph{Reshape the target if GPT proposes a smaller one.}
The deliberation's headline reshape\,---\,from the direct form
(\texttt{Candidate A}) to the centred form (X)\,---\,was load-bearing.
A direct-form proof would have had the
\texttt{(1/2)\,$\cdot\,$24$^{1/4}\,\cdot\,\Gamma$(1/4)} constant
threading through the DCT step; the centred form puts ``zero on the
right'' in the headline and lets the constant fall out by
\texttt{quartic\_partition} in a clean corollary. Cleaner statement,
cleaner proof, smaller line count. GPT correctly identified this as
a reshape worth the rename.

\item \emph{The K1 worktree discipline pays off immediately.}
Today this Tide ran with four other concurrent Tide branches active.
Zero collisions, zero file disappearances, zero stray-message
commits. The K1 wrapper is a structural fix to the failure class
documented in two prior retrospectives; this is the first Tide that
ran fully under it. Worth promoting in the Tide skill text from
``recommended'' to ``mandatory'' (already done).

\item \emph{Use \texttt{Continuous} when \texttt{AEMeasurable}
costs a Tide-substantial detour.} The deliberation correctly
identified the mathematically-minimal hypothesis class. The
formalisation correctly identified that the weakening is a
follow-up, not a Tide blocker. Both are right; the discipline is
to write \emph{both} into the tide log so the deliberation's intent
isn't lost.

\item \emph{Three new \texttt{CLAUDE.md} entries.} The `π`-as-binder
issue, the namespace inconsistency on \texttt{tendsto\_rpow\_atTop},
and the \texttt{∀ᶠ.symm} dot-notation gotcha are all small but
recurrent friction points. Worth landing.
\end{itemize}

\section{Follow-ups}

\begin{itemize}[leftmargin=*]
\item \emph{Hypothesis weakening to \texttt{AEMeasurable + ContinuousAt 0}.}
The deliberation's chosen minimum. Requires \texttt{AEMeasurable.comp\_quasiMeasurePreserving}
for the composition $\pi \circ (\cdot/a)$ plus a parallel argument
inside the DCT bound step. ~50 lines.

\item \emph{Generic-$k$ J-function for $K(w) = w^{2k}/(2k)!$.}
Mirror today's proof for $k=3$ (sextic, using \texttt{Sextic.lean}'s
\texttt{sextic\_partition}) and abstract a generic-$k$ template.
The \texttt{Real.rpow} exponent bookkeeping for $t^{1/(2k)}$ is the
main friction. Estimated ~250 lines as a follow-on Tide. Pairs with
the pending A1 (generic even-monomial 1D template) on the May 7
candidates survey.

\item \emph{Asymptotic-equivalence form $J(t) \sim \pi(0)\, Z(t)$.}
The grammar paper's actual statement is in equivalence form, not
\texttt{Tendsto}-times-power form. Conversion is one
\texttt{Tendsto.div}-style step away (\texttt{Tendsto (J(t)/Z(t)) atTop (𝓝 (\pi 0))})
once \texttt{Z(t)} is wrapped as a continuous nonzero function of
$t$. ~10 lines.

\item \emph{Promote the substitution lemma to \texttt{lean-common}.}
The pattern \emph{$t^{1/(2k)} \cdot \int f(t \cdot w^{2k}/(2k)!)\,dw =
\int f(u^{2k}/(2k)!)\,du$ for any locally bounded $f$} is generic to
single-monomial Laplace-method computations. Worth lifting to
\texttt{lean-common} once a second Laplace-method tide uses it.

\item \emph{\texttt{\textbackslash leanref}-tag the grammar paper.}
Once the grammar paper has a clear blueprint pointing at this
identity, tag with \texttt{\textbackslash leanref} pinned to commit
\texttt{5d2ae2c}. Process tide.

\item \emph{Mark E4 ✓ Closed in the May 7 candidates survey.}
SRI-side close-out per the Tide skill's Step-4 discipline.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step~2,
including the load-bearing reshape from Candidate~A (direct form)
to Candidate~X (centred form), the warning about the constant in
the generic-$k$ version (\texttt{1/k}, not \texttt{1/(2k)}), and the
recommendation against an \texttt{IsLittleO} reformulation. The full
consult is preserved at
\texttt{projects/primer/tide-log/gpt55\_tide\_grammar\_j\_function\_v1.md}.

\end{document}
