\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{∫}{\ensuremath{\int}}
\newunicodechar{∀}{\ensuremath{\forall}}
\newunicodechar{∃}{\ensuremath{\exists}}
\newunicodechar{ᶠ}{\ensuremath{{}^{f}}}
\newunicodechar{≤}{\ensuremath{\le}}
\newunicodechar{≥}{\ensuremath{\ge}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{→}{\ensuremath{\to}}
\newunicodechar{↦}{\ensuremath{\mapsto}}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{π}{\ensuremath{\pi}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{𝓝}{\ensuremath{\mathcal{N}}}

\title{Tide retrospective:\\\emph{Sextic J-function asymptotic\\(grammar precursor, Step 3)}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{2026-05-07}

\begin{document}
\maketitle

\begin{abstract}
For the sextic potential $K(w) = w^6/720$ and a continuous, globally
bounded prior $\pi : \mathbb{R} \to \mathbb{R}$, the J-function
$J(t) := \int_{\mathbb R} e^{-tK(w)}\,\pi(w)\,dw$ satisfies
$t^{1/6}\cdot J(t) \to \pi(0)\cdot \tfrac13\cdot 720^{1/6}\cdot \Gamma(1/6)$
as $t\to\infty$. This is the $k=3$ mirror of the $k=2$ quartic theorem
landed earlier today (E4 / \texttt{tide/grammar-j-function}) and the
third step of the grammar-paper precursor sequence. Two theorems in a
new file \texttt{Laplace/OneD/JFunctionSextic.lean}: the centred DCT
core \texttt{sextic\_jfunction\_centered\_tendsto\_zero} and the
user-facing asymptotic \texttt{sextic\_jfunction\_asymptotic} as a
two-line corollary via \texttt{sextic\_partition}. 280 lines of
Lean, $0$ \texttt{sorry}, $0$ \texttt{axiom}, $0$ \texttt{native\_decide}.
Single-session run; \texttt{lake build} green on first compile; no
GPT-5.5 Pro consult (the credentials file was missing in this
session, and Candidate~A was a direct mirror of E4 with the closed
form numerically verified to machine precision).
\end{abstract}

\section{Setting}

The grammar paper (Murfet \& Lehalleur, in preparation) needs a
catalogue of single-component blow-up J-function asymptotics. Step~1
of the precursor sequence (Tide~7) closed the quartic bounded-prior
partition. Step~2 (Tide~11+ / E4, today) closed the J-function
asymptotic for the quartic potential $K(w) = w^4/24$ at the full-line
scale, with a continuous-at-zero bounded prior. Step~3 — this Tide —
mirrors Step~2 to $K(w) = w^6/720$, the next monomial in the
sequence.

The seabed was \texttt{laplace} at \texttt{main} (commit
\texttt{141997c}), which already carried \texttt{Sextic.lean}'s
closed-form partition, integrability, and moment lemmas:
\[
  \int_{\mathbb R} e^{-t\cdot x^6/720}\, dx
    = \tfrac13 \cdot \big(\tfrac{720}{t}\big)^{1/6} \cdot \Gamma\!\big(\tfrac16\big),
\]
plus \texttt{sextic\_integrable\_pow} (polynomial-times-Gibbs
integrability for any $n\in\mathbb N$). E4's
\texttt{JFunctionQuartic.lean} was on the unmerged sibling branch
\texttt{tide/grammar-j-function} but visible via \texttt{git show};
its proof template — substitute, then DCT — was the natural mirror
target. The tide deliberation (\texttt{tide-pick}'s v2 survey of
2026-05-07) ranked Candidate~L1 as the obvious top pick: cache-fresh
after E4, single specific potential, single specific normalisation,
and the only friction is constants ($4 \to 6$, $24 \to 720$,
$1/4\to 1/6$, $1/2\to 1/3$).

The constant ratio is worth a sentence. The quartic case yielded
$\tfrac12\cdot 24^{1/4}\cdot \Gamma(1/4)$; the sextic yields
$\tfrac13\cdot 720^{1/6}\cdot \Gamma(1/6)$. The leading
$\tfrac1p$-vs-$\tfrac1{p}$ shift comes from Mathlib's
\texttt{integral\_rpow\_mul\_exp\_neg\_mul\_rpow}, which gives the
half-line integral with a factor $1/p$ for $K = b\, x^p$; the
full-line partition then doubles to $2/p$ ($\tfrac12$ for $p=4$,
$\tfrac13$ for $p=6$). Numerical sanity at $\pi=1$ matched the
closed form to $\sim 10^{-15}$ at every $t$ in the
$0.1$--$1000$ range.

\section{The thing formalised}

\begin{quote}\itshape
For a continuous prior $\pi : \mathbb R \to \mathbb R$ with global
bound $|\pi(w)| \le M$,
\[
  \lim_{t \to \infty}\, t^{1/6} \cdot \int_{\mathbb R} e^{-t\, w^6/720}\,\pi(w)\,dw
    \;=\;
  \pi(0) \cdot \tfrac13 \cdot 720^{1/6} \cdot \Gamma\!\big(\tfrac16\big).
\]
\end{quote}

\noindent The Lean signature
(\texttt{Laplace.OneD.sextic\_jfunction\_asymptotic}):
\begin{verbatim}
theorem sextic_jfunction_asymptotic
    {prior : ℝ → ℝ} (hprior_cont : Continuous prior)
    {M : ℝ} (hprior_bd : ∀ x, |prior x| ≤ M) :
    Tendsto (fun t : ℝ =>
        t ^ ((1:ℝ)/6) * ∫ w : ℝ, Real.exp (-(t * w^6 / 720)) * prior w)
      atTop (𝓝 (prior 0 * (1/3) * (720:ℝ) ^ ((1:ℝ)/6) * Real.Gamma (1/6)))
\end{verbatim}

The load-bearing theorem is the centred form
\texttt{sextic\_jfunction\_centered\_tendsto\_zero}, which states that
$t^{1/6}\cdot \int e^{-tw^6/720}\,(\pi(w)-\pi(0))\,dw \to 0$. This is
the form the grammar paper will cite in its J-function asymptotic
identity: the prior perturbation $\pi(w) - \pi(0)$ is asymptotically
negligible against the concentrating Gibbs weight, and so the
J-function is determined to leading order by $\pi(0)$ alone. The
asymptotic-form corollary above is a two-line wrapper that splits
$\pi(w) = (\pi(w) - \pi(0)) + \pi(0)$, applies the centred theorem to
the perturbation, and applies \texttt{sextic\_partition} (closed-form)
to the constant piece, with the cancellation
$t^{1/6}\cdot (720/t)^{1/6} = 720^{1/6}$ tying it together.

The hypotheses match E4: \texttt{Continuous prior} plus a global
$L^\infty$ bound. The mathematically minimal class is
\texttt{AEMeasurable + ContinuousAt prior 0 + global bound}; we
deliberately did \emph{not} weaken in this tide, so the sextic and
quartic files share an API surface. The hypothesis-weakening is E4
follow-up~L3 and should be done as a single refactor across both
files (and a \texttt{lean-common} substitution lemma) — landing it
asymmetrically here would have created an awkward sextic-stronger
quartic-weaker mismatch.

\section{Proof strategy}

The proof is substitute-then-DCT, identical in structure to E4 with
the obvious constant rewrites.

\paragraph{The substitution.} Set $a := t^{1/6}$, so $a^6 = t$. The
change of variables $u = a\cdot w$ via Mathlib's
\texttt{Measure.integral\_comp\_mul\_right} reshapes
\[
  t^{1/6} \int_{\mathbb R} e^{-t w^6/720}\,(\pi(w) - \pi(0))\,dw
\]
into
\[
  \int_{\mathbb R} e^{-u^6/720}\,(\pi(u/a) - \pi(0))\,du.
\]
The Jacobian bookkeeping ($a > 0$, $|a^{-1}| = 1/a$,
$(w\cdot a)^6 = t\, w^6$, $(w\cdot a)/a = w$) is the load-bearing
piece — about $50$ of the file's $280$ lines. The cast
$a^{(6:\mathbb N)} = a^{(6:\mathbb N : \mathbb R)}$ uses
\texttt{Real.rpow\_natCast}; the exponent identity
$a^6 = (t^{1/6})^6 = t^{(1/6)\cdot 6} = t^1 = t$ uses
\texttt{Real.rpow\_mul} and \texttt{Real.rpow\_one}.

\paragraph{The DCT.} The substituted integrand
$F(t,u) := e^{-u^6/720}\cdot (\pi(u/a) - \pi(0))$ has a
$t$-independent envelope $\mathrm{bound}(u) := 2M\cdot e^{-u^6/720}$,
integrable from \texttt{sextic\_integrable\_pow}~$0$ at~$t=1$. The
pointwise limit is zero: as $t \to \infty$,
$u/t^{1/6} \to 0$ (from \texttt{tendsto\_rpow\_atTop} at exponent
$1/6 > 0$), and continuity of $\pi$ at~$0$ gives
$\pi(u/t^{1/6}) - \pi(0) \to 0$. Mathlib's
\texttt{tendsto\_integral\_filter\_of\_dominated\_convergence}
closes.

\paragraph{The asymptotic corollary.} Split
$\pi(w) = (\pi(w) - \pi(0)) + \pi(0)$ inside the integrand. The
centred piece tends to zero by the headline theorem. The constant
piece equals $\pi(0)\cdot \int e^{-tw^6/720}\,dw$, which by
\texttt{sextic\_partition} is $\pi(0)\cdot \tfrac13 \cdot
(720/t)^{1/6}\cdot\Gamma(1/6)$. Multiplying by $t^{1/6}$, the
$(720/t)^{1/6}$ factor combines with $t^{1/6}$ via
$t^{1/6}\cdot (720/t)^{1/6} = 720^{1/6}$ (one
\texttt{Real.mul\_rpow} step plus \texttt{field\_simp}).

The whole proof is therefore a two-step move: a $50$-line substitution
identity, then a four-clause invocation of DCT, then an integration-
splitting corollary. No mathematical novelty over E4; the whole tide
is a constants-substitution exercise, exactly as a Step-3-of-the-
sequence tide should be.

\section{Roadblocks and resolutions}

There were no serious roadblocks. The file compiled green on first
compile after the \texttt{lake exe cache get} step. The substitution
identity, which absorbed most of E4's effort, was a direct port —
swap $\tfrac14$ for $\tfrac16$ throughout, swap $a^4$ for $a^6$, swap
\texttt{quarticPotential} for \texttt{sexticPotential}, and the
\texttt{ring} / \texttt{field\_simp} closures all fire identically.

The only friction was infrastructural: the worktree was created via
the K1 worktree wrapper but did not inherit the canonical clone's
\texttt{.lake/build/} directory, so the first \texttt{lake build}
attempt started compiling Mathlib from source. Killed and re-ran with
\texttt{lake exe cache get} first, which restored the prebuilt oleans
in $\sim$15 seconds; the subsequent build took $\sim$2 minutes
(target file alone in $3.1$~s). This is a one-line addition to the
worktree wrapper: every fresh worktree should run
\texttt{lake exe cache get} as part of \texttt{tide-worktree create}.

A second infrastructural friction: the GPT-5.5 Pro credentials file
at \texttt{.agents/skills/timaeus-research/credentials.json} was
missing from the session. Earlier today's tides had access; somehow
the file was removed or never copied locally. The deliberation
proceeded on Claude judgment alone, justified by the fact that the
candidate was a direct mirror of an already-deliberated proof and
the closed-form constant was numerically verified at machine
precision. This is acceptable for mirror-shaped tides; for any tide
with substantive mathematical novelty, the credentials wall would
need to be unblocked first.

\section{What was Mathlib, what was new}

Mathlib provided the entire infrastructure. Specifically:
\texttt{MeasureTheory.Measure.integral\_comp\_mul\_right} for the
substitution; \texttt{Real.rpow\_natCast},
\texttt{Real.rpow\_mul}, and \texttt{Real.rpow\_one} for the exponent
arithmetic; \texttt{Real.mul\_rpow} for the
$t^{1/6}\cdot(720/t)^{1/6} = 720^{1/6}$ cancellation;
\texttt{tendsto\_rpow\_atTop} for $t^{1/6} \to \infty$;
\texttt{MeasureTheory.tendsto\_integral\_filter\_of\_dominated\_convergence}
for the DCT; \texttt{MeasureTheory.integral\_add},
\texttt{MeasureTheory.integral\_mul\_const}, and
\texttt{Integrable.const\_mul} / \texttt{Integrable.mul\_const} for
the linearity bookkeeping.

What was new was, modulo constants, identical to what was new in
E4: the substitution lemma
\texttt{sextic\_jfunction\_centered\_subst}, the bounded-dominator
wrapper \texttt{sextic\_dominator\_integrable}, the DCT theorem, the
asymptotic corollary. No genuinely new \emph{mathematical} content
appeared; the tide's contribution is to extend the closed-form
J-function template from $k=2$ to $k=3$ inside the same Lean
infrastructure, which is the small step the grammar-precursor
sequence needs to demonstrate the template generalises.

\section{Lessons}

\begin{itemize}[leftmargin=1em,itemsep=0.2em]
\item \emph{Worktree wrapper should run \texttt{lake exe cache get}.}
The K1 wrapper creates a worktree from \texttt{git worktree add}
but does not seed \texttt{.lake/build/}. The fresh worktree has
\texttt{.lake/packages/mathlib/.lake/build/} only partially
populated (whatever was synced as part of clone), and a naive
\texttt{lake build} re-compiles ${\sim}$2700 Mathlib files. The
fix is one line: append \texttt{lake exe cache get} to
\texttt{tide-worktree create}'s tail. (Promote to the
\texttt{tide} skill's branch hygiene preflight too.)

\item \emph{Credentials hygiene.} The
\texttt{.agents/skills/timaeus-research/credentials.json} file
should be a session-startup invariant for any tide that may
need GPT. Worth adding a one-line preflight check to the
\texttt{tide} skill: if the credentials are missing, surface to
the user before Step~2.

\item \emph{Mirror tides are cheap.} A direct constants-mirror of an
already-deliberated tide can land in a single short session with no
GPT consult, provided the closed form is numerically verified at
the destination constants. Cost ratio for this tide vs E4
(both single-session, similar friction): roughly $1{:}3$ in
agent-time, attributable entirely to the absence of
deliberation overhead and the inherited proof template.

\item \emph{The $1/p$-vs-$2/p$ partition prefactor needs flagging.}
The constant differs between $p=4$ and $p=6$ ($\tfrac12$
vs.~$\tfrac13$), and a copy-paste agent that mirrors E4
mechanically without checking \texttt{sextic\_partition}'s actual
right-hand side will land the wrong constant. The v2 survey did
exactly this in its L1 entry (it copied $\tfrac12$ from E4); the
tide-log corrected the survey before formalising. Worth a flag
in the \texttt{tide-pick} skill: when proposing a mirror
candidate, surface the destination's closed-form partition
explicitly so the survey doesn't carry a copy-paste constant.
\end{itemize}

\section{Follow-ups}

\begin{itemize}[leftmargin=1em,itemsep=0.2em]
\item \emph{Generic-$k$ abstraction (E4 follow-up L1, now ripe).} With
both $k=2$ and $k=3$ in the file and identical proof structure,
the abstraction over $K(w) = w^{2k}/(2k)!$ is the natural next
step. Closed-form constant
$\tfrac{1}{k}\cdot ((2k)!)^{1/(2k)}\cdot \Gamma(1/(2k))$, modulo
factors. Estimated ${\sim}$300 lines for the parametric form.

\item \emph{Asymptotic-equivalence wrapper $J(t) \sim \pi(0)\cdot Z(t)$
(E4 follow-up L2).} Now applies to two files (quartic, sextic).
Probably worth a single shared $\sim$15-line corollary that takes
the \texttt{Tendsto (t\textsuperscript{1/p}\cdot J)} form and
divides through by the partition function.

\item \emph{Hypothesis weakening (E4 follow-up L3).} Defer until both
quartic and sextic land on \texttt{main}; then refactor both
files together to use \texttt{AEMeasurable + ContinuousAt 0 +
global bound}, with a shared substitution lemma in
\texttt{lean-common}.

\item \emph{Promote \texttt{jfunction\_centered\_subst} to
\texttt{lean-common} (E4 follow-up L4).} Would shrink both
quartic and sextic files by ${\sim}50$ lines each. Best done
together with the L3 hypothesis-weakening so both files are
visited once.

\item \emph{Adopt the worktree-cache-seeding fix (this tide's lesson).}
One-line addition to \texttt{tide-worktree create}; expect to
land as a process tide.

\item \emph{\textbackslash leanref-tag the grammar paper (E4 follow-up
L5).} Once the grammar paper has TeX statements pinning the
quartic and sextic J-function asymptotics, tag both with
\texttt{\textbackslash leanref} pinned to commit
\texttt{be4608f} (this tide).
\end{itemize}

\end{document}
