\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{²}{\ensuremath{^2}}
\newunicodechar{⁶}{\ensuremath{^6}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{√}{\ensuremath{\sqrt{\,}}}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≥}{\ensuremath{\geq}}
\newunicodechar{≠}{\ensuremath{\neq}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{Sextic 1D moments}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{2 May 2026}

\begin{document}
\maketitle

\begin{abstract}
A single Tide-loop excursion in the \texttt{laplace} seabed, generalising
the pure-quartic 1D file from $L(x) = x^4/24$ to $L(x) = x^6/720$. The
headline theorem (\texttt{sextic\_cov\_affine}) gives the affine covariance
of the 1D pure-sextic Gibbs measure as
$\operatorname{Cov}_t[ax+c, bx+d] = ab \cdot (720/t)^{1/3} \cdot
\Gamma(1/2)/\Gamma(1/6)$, a rate-$t^{-1/3}$ analogue of the quartic's
$t^{-1/2}$. The new file (396 lines) parallels its quartic predecessor
section by section. Zero \texttt{sorry}, zero \texttt{axiom}, zero
\texttt{native\_decide}; clean Lean diagnostics on first pass. The
interesting features of this run were two tactical refinements suggested by
GPT-5.5 Pro at the deliberation step --- a \emph{B+ refinement} stating the
half-line moment lemma for arbitrary natural power $m$ rather than just
$2n$, and a piecewise (rather than discriminant-based) proof of the
polynomial bound $x^2 \leq 1 + x^6$ used for Gaussian-comparison
integrability --- both of which made the file genuinely cleaner than the
quartic template it descended from. Like the concurrent 2D pure-quartic
excursion, this run hit a working-tree race when another agent rewrote the
top-level import file in the staging window; the fix was atomic re-staging
plus a branch-ref re-pointing dance.
\end{abstract}

\section{Setting}

This Tide step is the sixth excursion outward from the same seabed: the
\texttt{laplace} repo, which originally formalised the 1D anharmonic
asymptotic $\operatorname{Cov}_t[x^2,x] = -2\alpha/(\lambda^3 t^2) +
o(t^{-2})$ (Susceptibility Primer, eq.~4.10) and has since been extended
through five prior tide steps.\footnote{Branches
\texttt{tide/quartic-moments}, \texttt{tide/2d-semi-degenerate},
\texttt{tide/universal-scaling}, \texttt{tide/f1-large-t}, and the
concurrent \texttt{tide/2d-pure-quartic}.}
The first four merged into \texttt{main} on 1\,May (commit
\texttt{44a6001}); this step shares its base with the parallel 2D
pure-quartic run and was branched off the latter's tip per the Tide skill's
branch-hygiene rule.

The candidate was picked from a cross-tide candidate survey written
earlier the same day.\footnote{\texttt{projects/automation/log/}\allowbreak\texttt{2026-05-02-tide-candidates-survey.md}.}
Of the five candidates listed there, \emph{Sextic 1D moments} scored as the
third choice, behind the threepoint $\kappa_3$ candidate (different seabed)
and the 2D pure-quartic case (the parallel run). It is the cleanest test
of whether the substitution-and-$\Gamma$ template established by the
quartic file generalises beyond a single exponent class. The pattern
$L = x^{2k}/(2k)!$ predicts a rate $t^{-1/(2k-1)}$, with $k=2$ giving the
quartic's $t^{-1/2}$ and $k=3$ giving the sextic's $t^{-1/3}$; this tide is
the second data point.

The Step~2 deliberation (Claude $\leftrightarrow$ GPT-5.5 Pro) drafted three
candidates: A, the full sextic package paralleling the quartic file; B,
just the atomic half-line moment as a kernel; C, the general $p = 2k$
family covering quartic and sextic uniformly. Both parties backed A, with
GPT making two tactical refinements that were carried into Step~3. The full
deliberation is in the tide log.\footnote{\texttt{projects/primer/tide-log/}\allowbreak\texttt{2026-05-02-tide-sextic-moments.md}.}

\section{The thing formalised}

For the centred sextic 1D potential
\[
  L(x) = \frac{x^6}{720}
\]
and affine observables $\varphi(x) = ax + c$, $\psi(x) = bx + d$, against the
1D Gibbs measure $e^{-tL(x)}\,dx / Z(t)$:

\begin{quote}\itshape
For all $t > 0$,
\[
  \operatorname{Cov}_t[\varphi,\psi]
  \;=\; ab \cdot \left(\frac{720}{t}\right)^{1/3} \cdot
        \frac{\Gamma(1/2)}{\Gamma(1/6)}
  \;=\; ab \cdot \frac{\sqrt{\pi}\,(720/t)^{1/3}}{\Gamma(1/6)}.
\]
\end{quote}

\noindent The Lean signature:
\begin{verbatim}
theorem sextic_cov_affine {t : ℝ} (ht : 0 < t) (a b c d : ℝ) :
    gibbsCov sexticPotential t (fun x => a * x + c) (fun x => b * x + d) =
      a * b * (720 / t) ^ ((1 : ℝ) / 3) * Real.Gamma ((1 : ℝ) / 2) /
        Real.Gamma ((1 : ℝ) / 6)
\end{verbatim}

\noindent located at \texttt{Laplace/OneD/Sextic.lean}, in namespace
\texttt{Laplace.OneD}. The file commits twelve further lemmas in addition
to the headline: the potential definition, two flavours of integrability,
the atomic half-line moment for arbitrary natural power, even and odd
full-line moments, the partition function and its positivity, even and odd
expected values, and the $n=1$ specialisation
$\langle x^2\rangle_t = (720/t)^{1/3}\Gamma(1/2)/\Gamma(1/6)$ which is the
load-bearing input to \texttt{sextic\_cov\_affine}.

The shape $L(x) = x^{2k}/(2k)!$ means that the unique critical point
$x = 0$ has $L^{(j)}(0) = 0$ for $j = 0, \dots, 2k-1$ and
$L^{(2k)}(0) = 1$: this is the rank-zero degenerate case,
analogous to the quartic ($k=2$) but more degenerate. The Hessian-based
Laplace machinery from the rest of the seabed does not apply; instead the
substitution $u = (t/720)^{1/6} x$ reduces every moment integral exactly
to a $\Gamma$-function value, and the rate $t^{-1/3}$ falls out of the
$1/(2k-1)$ formula at $k=3$.

Numerical sanity check before formalising: \texttt{scipy.integrate.quad}
agreed with the closed forms to $\sim 10^{-12}$ relative error across
$t \in \{0.1, 1.0, 10.0, 100.0\}$ and $n \in \{0,1,2,3\}$, for the half-line
moments, the full-line even moments, the partition function, the expected
values, and the affine covariance. The bonus integration-by-parts identity
$\langle x^6\rangle_t = 120/t$ (independent of all the special-function
constants) provides a Gibbs-side cross-check that is also clean.

\section{Proof strategy}

The atomic kernel of the file is the half-line moment
\texttt{integral\_pow\_mul\_exp\_neg\_sextic\_Ioi}, stated for arbitrary
natural power $m$:
\[
  \int_0^\infty x^m e^{-t x^6 / 720}\,dx
  \;=\; \tfrac{1}{6} \left(\tfrac{720}{t}\right)^{(m+1)/6}
        \Gamma\!\left(\tfrac{m+1}{6}\right).
\]
This is a direct application of Mathlib's master gamma-integral
lemma\footnote{\texttt{integral\_rpow\_mul\_exp\_neg\_mul\_rpow}.} with
$p = 6$, $q = m$, $b = t/720$, plus a conversion between $\mathbb{N}$- and
$\mathbb{R}$-powers via \texttt{rpow\_natCast} and a single
$(t/720)^{-(m+1)/6} \to (720/t)^{(m+1)/6}$ inversion via \texttt{inv\_rpow}.
Stating this for all $m$ (the \emph{B+ refinement}) costs nothing extra in
the proof and makes the even and odd full-line cases share a single
substitution argument.

From the half-line lemma the rest of the file is mechanical. The full-line
even moment doubles via \texttt{integral\_comp\_abs} after rewriting the
integrand in $|x|$-form (this uses
$x^{2n} = |x|^{2n}$ and $x^6 = |x|^6$, both via \texttt{sq\_abs}); the odd
moment vanishes via the same parity argument the quartic file used,
\texttt{Odd.neg\_pow $\langle n,\text{rfl}\rangle$} together with
$(-x)^6 = x^6$ to flip the sign of the integrand. The partition function is
the $n=0$ specialisation of the even moment, the expected value is the
ratio of the moment by the partition function (with the cancellation
$(720/t)^{(2n+1)/6 - 1/6} = (720/t)^{n/3}$), and the affine covariance
expands $(ax+c)(bx+d)$, uses linearity of \texttt{gibbsExpectation} (which
itself is unfolded against \texttt{integral\_add} and
\texttt{integral\_const\_mul} after building integrability witnesses for
$1, x, x^2$ against the Gibbs weight), and collapses via parity of the
$\langle x\rangle$ terms to $ab \langle x^2\rangle_t$.

The integrability of $x^n e^{-t x^6/720}$ against Lebesgue measure is
the only place the proof has any choices. The quartic file used an exact
completing-the-square identity
$t \cdot x^4/24 - x^2 = (tx^2 - 12)^2/(24t) - 6/t$
to dominate by a Gaussian with $b = 1$ and prefactor $e^{6/t}$. The sextic
analogue would need to minimise a cubic $t \cdot x^6/720 - x^2$, which is
messy in Lean. The route taken here, on GPT-5.5 Pro's tactical advice, is
the slick polynomial bound
\[
  x^2 \;\leq\; 1 + x^6
\]
proved piecewise: on $|x| \leq 1$, $x^2 \leq 1 \leq 1 + x^6$; on
$|x| \geq 1$, $x^4 \geq 1$ so $x^2 \leq x^2 \cdot x^4 = x^6 \leq 1 + x^6$.
Multiplying through by $t/720 > 0$ gives
$t \cdot x^6/720 \geq (t/720) x^2 - t/720$, hence
$e^{-t x^6/720} \leq e^{t/720} \cdot e^{-(t/720) x^2}$, dominated by a
Gaussian of width $t/720$. Mathlib's
\texttt{integrable\_rpow\_mul\_exp\_neg\_mul\_sq} handles the dominator
directly with $b = t/720$; the rest is bookkeeping. The piecewise lemma is
nine lines; its quartic counterpart was thirteen.

\section{Roadblocks and resolutions}

\paragraph{None of mathematical substance.} The proof typechecked clean on
the first compile after a one-line tactic fix
(\texttt{rcases le\_or\_lt} $\to$ \texttt{rcases le\_total}; \texttt{le\_or\_lt}
no longer exists in the version of Mathlib pinned by the seabed) and a
one-line cleanup of an unused \texttt{push\_cast} that had been a vestige of
copy-pasting from the quartic file. The candidate had been picked precisely
so that the seabed's machinery would already cover the moves; the run paid
the cost of that prediction.

\paragraph{Concurrent-agent staging race.} Same root cause as the parallel
2D pure-quartic Tide. Multiple times during this run, the working-tree copy
of \texttt{Laplace.lean} was overwritten by the other agent who was also
adding an import for their new file. The interesting variant, in this run,
was that the \texttt{git commit} fired while the working tree was on the
\emph{wrong branch}: at the moment of commit, the parallel agent had checked
out their \texttt{tide/2d-pure-quartic} branch in the shared working tree, so
my Sextic commit landed on \emph{their} branch rather than my own. The fix
was

\begin{verbatim}
git branch -f tide/sextic-moments HEAD       # claim the commit
git checkout tide/sextic-moments
git branch -f tide/2d-pure-quartic HEAD~1    # restore their tip
\end{verbatim}

\noindent which by happy accident produced exactly the linear-stack
arrangement the Tide skill's branch-hygiene rule requires (Sextic on top of
PureQuartic on top of \texttt{main}), but it was a recovery move rather than
a deliberate one. After the parallel run later added a trailing
\texttt{Laplace.lean} import-list fix-up commit and a retrospective commit
on top of their original tip, this branch was rebased onto the new tip via
\texttt{git rebase tide/2d-pure-quartic}, which applied without conflict.

The structural lesson is the same as the parallel retrospective concluded:
concurrent Tide agents need filesystem isolation. \texttt{git worktree} per
Tide branch is the right answer.

\paragraph{Tactical advice from GPT-5.5 Pro that paid off.} Two pieces of
GPT-5.5 Pro's deliberation-stage response were carried into the proof and
each paid for itself.

\begin{itemize}\itemsep0pt

\item \emph{The B+ refinement.} GPT noted that on $\mathrm{Ioi}\,0$, parity is
irrelevant, so stating the half-line moment for arbitrary natural power $m$
costs nothing extra and makes the file's even/odd full-line story share a
single substitution argument. This is the kind of strict-generalisation-for-free
that Tide-loop deliberation is for; carrying it in saved a near-duplicate
half-line argument for the odd case.

\item \emph{Piecewise integrability.} GPT pre-empted the obvious sin (which
I had drafted) of proving $x^2 \leq 1 + x^6$ via a cubic-discriminant
argument:

\gptbox{Your proposed bound $x^2 \leq 1 + x^6$ is excellent, but in Lean I
would \textbf{not} prove it via discriminant/minimum. I would prove it
piecewise: if $|x| \leq 1$, then $x^2 \leq 1 \leq 1 + x^6$; if
$1 \leq |x|$, then $x^2 \leq x^6$, hence $x^2 \leq 1 + x^6$. That will be
much friendlier than a cubic-minimum argument.}

\noindent The piecewise version that landed is nine lines, three of which
are \texttt{nlinarith}-with-hint closers; the discriminant version would
have been a forty-line minimisation by hand. This was the highest-value
tactical advice in the deliberation.
\end{itemize}

Beyond these two, no proof-side roadblocks: the file is essentially a
substitute-and-rename of \texttt{Quartic.lean} with the integrability lemma
swapped out and the half-line lemma generalised in $m$.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.}
\texttt{integral\_rpow\_mul\_exp\_neg\_mul\_rpow} for the master half-line
gamma integral. \texttt{integrable\_rpow\_mul\_exp\_neg\_mul\_sq} for the
Gaussian dominator. \texttt{integral\_comp\_abs} for the
even-symmetric reduction to the half line. \texttt{integral\_neg\_eq\_self}
for the odd parity argument. The standard
\texttt{integral\_add\,/\,integral\_const\_mul\,/\,Integrable.const\_mul/add}
lemmas for the affine-observable expansion. \texttt{rpow\_natCast},
\texttt{inv\_rpow}, \texttt{rpow\_neg}, \texttt{Real.rpow\_add} for exponent
algebra. \texttt{Real.Gamma\_pos\_of\_pos} for partition-function positivity.

\paragraph{The seabed provided.}
\texttt{Laplace.Gibbs} (\texttt{partitionFunction}, \texttt{gibbsExpectation},
\texttt{gibbsCov} for the 1D Gibbs interface), and
\texttt{Laplace/OneD/Quartic.lean} as the structural template. No new
seabed-side infrastructure was needed.

\paragraph{What was new in this tide.}
The 1D pure-sextic potential, a Gaussian-comparison integrability lemma
(with the piecewise polynomial bound $x^2 \leq 1 + x^6$ as a private
helper), the half-line moment lemma in the B+ form for arbitrary natural
power, the full-line even and odd moments, the partition function and its
positivity, the four expected-value lemmas (even, odd, $n=1$, linear
specialisation), and the headline affine
covariance.\footnote{Lean names: \texttt{sexticPotential};
\texttt{sextic\_integrable\_pow}; \texttt{sq\_le\_one\_add\_pow\_six} (private);
\texttt{integral\_pow\_mul\_exp\_neg\_sextic\_Ioi};
\texttt{sextic\_moment\_even}, \texttt{sextic\_moment\_odd};
\texttt{sextic\_partition}, \texttt{sextic\_partition\_pos};
\texttt{sextic\_expected\_value\_even}, \texttt{\_odd},
\texttt{\_sq}, \texttt{\_lin};
\texttt{sextic\_cov\_affine}.}
There is no novel mathematical content; what is new is the
substitution-into-the-template that confirms the substitution template
ports cleanly to a second exponent class. As the parallel 2D pure-quartic
retrospective predicted in its \emph{Lessons}: ``the next test is whether
[the factorisation template] ports across the sextic side; if so, the
abstraction is genuinely the right shape.'' It does, and it is.

\section{Lessons}

\begin{itemize}
\item \emph{Specific tactical advice from GPT-5.5 Pro at the deliberation
stage transfers cleanly into Step~3.} The B+ refinement and the piecewise
integrability bound were both proposed at Step~2 and carried verbatim into
the final proof. This is the right division of labour: the deliberation
stage is where ``how you should phrase the lemma'' and ``how you should
prove the polynomial inequality'' get decided, before any Lean is written.
The cost was a single 5-minute consult; the value, fifty lines of avoided
proof.

\item \emph{The substitution-and-$\Gamma$ template is robust at $k=3$.} The
$L = x^{2k}/(2k)!$ family produces, at each $k$, a rate $t^{-1/(2k-1)}$
covariance with constants in $\Gamma((2n+1)/(2k))/\Gamma(1/(2k))$ and a
proof that mirrors the previous $k$ section by section. With $k=2$ and
$k=3$ both done in essentially the same shape, candidate C from the
deliberation (the unifying \texttt{p = 2k} parameterisation) becomes
substantially more attractive than it did at $k=2$ alone --- the case for
``just refactor and parameterise'' has gone from one data point to two.

\item \emph{Concurrent Tide agents in the same checkout fight over
\texttt{Laplace.lean}.} Same lesson as the parallel retrospective draws.
Filesystem isolation via \texttt{git worktree} is the right structural fix;
ad-hoc atomic re-staging is at best a recovery move.

\item \emph{Piecewise polynomial bounds beat discriminant-based ones in
Lean.} The \texttt{nlinarith}-with-hint closers do not like cubic minima,
but a case split on $|x| \leq 1$ vs $|x| \geq 1$ followed by a single
\texttt{nlinarith} per branch is robust. Worth lifting to the seabed's
\texttt{CLAUDE.md} as a tactic note.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{The \texttt{p = 2k} parameterisation, now plausible.} With
\texttt{Quartic.lean} ($k=2$) and \texttt{Sextic.lean} ($k=3$) now both in
place and structurally near-identical, the unifying lemma
\[
  \int_0^\infty x^m e^{-t x^{2k}/(2k)!}\,dx
  \;=\; \tfrac{1}{2k} \left(\tfrac{(2k)!}{t}\right)^{(m+1)/(2k)}
        \Gamma\!\left(\tfrac{m+1}{2k}\right)
\]
is now reachable in roughly the same proof body as the present file, just
with $k$ generic. Best opened as a future Tide step: not a refactor of the
existing files (they stay as worked examples), but a new
\texttt{Laplace/OneD/PolyEven.lean} (or similar) that proves the parameterised
identity once and recovers the quartic and sextic specialisations as
corollaries. Candidate B and C from the present deliberation would land
together as part of that step.

\item \emph{Quartic--sextic mixed potential.} $L(x,y) = x^4/24 + y^6/720$,
with rates $t^{-1/2}$ in $x$ and $t^{-1/3}$ in $y$. Mechanical given the
present file plus \texttt{Laplace/TwoD/PureQuartic.lean} from the parallel
Tide, particularly if the additively-separable factorisation refactor
(deferred in that retrospective) lands first.

\item \emph{Mathlib upstreaming for polynomial-times-Gibbs integrability.}
\texttt{quartic\_integrable\_pow} and \texttt{sextic\_integrable\_pow} are
the same shape: $\int x^n \exp(-b \cdot x^{2k})$ for $b > 0, k \geq 2,
n \in \mathbb{N}$. With two concrete uses, the case for promoting a generic
version to \texttt{lean-common} (or upstream into Mathlib alongside the
existing \texttt{integrable\_rpow\_mul\_exp\_neg\_mul\_sq}) is now stronger
than at one use.

\item \emph{\texttt{\textbackslash leanref} tagging.} If the primer gains a
sextic-rate paragraph in its degenerate-case discussion, tag it with
\texttt{\textbackslash leanref\{sextic\_cov\_affine\}}. Currently no such
paragraph exists; the sextic case is implicit in the
$L = x^{2k}/(2k)!$ generalisation that the primer alludes to but does not
spell out. Worth a one-paragraph appendix addition the next time the
primer's degenerate-case section is touched.

\item \emph{Worktree-isolated Tide runs.} Same as the
\texttt{tide/2d-pure-quartic} retrospective notes: a small wrapper that
creates a \texttt{git worktree} per Tide branch and runs the agent against
that worktree would prevent the staging race. Worth adding to the Tide
skill as a preflight setup step.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step~2,
including the B+ refinement of the half-line moment lemma and the piecewise
proof of the polynomial bound $x^2 \leq 1 + x^6$. Both were carried into
the final proof, and both were the right call. The full consult and the
tide log (with candidate drafts, vote, numerical check, and result) are
both preserved in the primer project's tide-log
directory.\footnote{\texttt{projects/primer/tide-log/}\allowbreak\texttt{gpt55\_tide\_sextic\_v1.md}
and
\texttt{projects/primer/tide-log/}\allowbreak\texttt{2026-05-02-tide-sextic-moments.md}.}
The parallel 2D pure-quartic run shared the seabed and the
integration race; its retrospective is the immediate predecessor to this
one and recorded the same lessons from the other side.

\end{document}
