\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{κ}{\ensuremath{\kappa}}

\title{Tide retrospective:\\third-moment asymptotic for the anharmonic Gibbs}
\author{Daniel Murfet}
\date{7 May 2026}

\begin{document}
\maketitle

\begin{abstract}
This retrospective records the smallest tide on the board: a $\sim$30-line
strict-improvement of Tide 9. The third moment $\langle x^3 \rangle_t$ of
the anharmonic Gibbs measure was internal to the proof of
\texttt{cov\_anharmonic\_asymptotic} (Tide 9); this tide packages it as a
separate user-facing theorem
$t^2 \cdot \langle x^3 \rangle_t \to -5\alpha/(2\lambda^3)$ via the
algebraic identity $\langle x^3 \rangle = \mathrm{Cov}[x^2, x] + \langle x^2 \rangle \langle x \rangle$
combined with the existing asymptotics. Zero \texttt{sorry}, single
session, no GPT-5.5~Pro consult needed (the deliberation already happened
in Tide 9's retrospective).
\end{abstract}

\section{Setting}

Tide 9 (anharmonic-$\kappa_3$, 6 May) proved
$t^2 \cdot \kappa_3(x, x, x) \to -\alpha/\lambda^3$ for the 1D anharmonic
Gibbs measure. The proof factored through three asymptotics already in
the laplace seabed:\footnote{Specifically, in
\texttt{Laplace/OneD/IntegralRemainder.lean}:
\texttt{cov\_anharmonic\_asymptotic} (the covariance),
\texttt{mean\_anharmonic\_asymptotic} ($\langle x \rangle$), and
\texttt{cov\_self\_anharmonic\_asymptotic} (the variance).}
the covariance, the mean, and the variance. Internal to the
strategy was the third moment expansion
$\langle x^3 \rangle_t = -5\alpha/(2\lambda^3 t^2) + o(t^{-2})$, which
appeared as the corrected Wick coefficient (the famous "5/2 vs 3/2"
correction caught by both the numerical sanity check and the GPT consult).

That third moment was never packaged as a stand-alone user-facing
theorem. Tide 9's retrospective explicitly named it as a deferred
follow-up: \emph{``With Tide 9's moment expansions exposed,
$t^2 \cdot \langle x^3 \rangle_t \to -5\alpha/(2\lambda^3)$ is a $\sim$5-line
corollary of $\mathrm{Cov}[x^2, x] = \langle x^3 \rangle - \langle x^2 \rangle \langle x \rangle$
plus existing asymptotics. Worth packaging as
\texttt{thirdMoment\_anharmonic\_asymptotic} for downstream consistency.''}

This tide closes that follow-up, exactly as Tide 9 anticipated.

\section{The thing formalised}

\begin{quote}\itshape
For the 1D anharmonic potential
$L(x) = (\lambda/2) x^2 + (\alpha/6) x^3 + (\gamma/24) x^4$ with
$\lambda > 0$, $\gamma > 0$, and the discriminant condition
$\alpha^2 < 3\lambda\gamma$,
\[
  t^2 \cdot \langle x^3 \rangle_t
  \;\longrightarrow\; -\frac{5\alpha}{2\lambda^3}
  \quad \text{as } t \to \infty.
\]
\end{quote}

The Lean signature, in
\texttt{Laplace/OneD/AnharmonicKappa3.lean}\footnote{The new theorem
sits at the bottom of that file; the existing $\kappa_3$ infrastructure
remains unchanged. Permalink:
the file at the merged commit on \texttt{main}.}:

\begin{verbatim}
theorem thirdMoment_anharmonic_asymptotic
    {lam alpha gamma : ℝ}
    (hlam : 0 < lam) (hgamma : 0 < gamma)
    (hdisc : alpha ^ 2 < 3 * lam * gamma) :
    Filter.Tendsto
      (fun t : ℝ => t ^ 2 * Laplace.gibbsExpectation
          (anharmonicPotential lam alpha gamma) t (fun x : ℝ => x ^ 3))
      Filter.atTop
      (nhds (-(5 * alpha) / (2 * lam ^ 3)))
\end{verbatim}

The hypotheses are inherited verbatim from Tide 9.

\section{Proof strategy}

The proof is short enough to write out completely in prose:

By the definition of \texttt{gibbsCov},
\[
  \mathrm{Cov}_t[x^2, x] = \langle x^3 \rangle_t - \langle x^2 \rangle_t \cdot \langle x \rangle_t,
\]
so
\[
  \langle x^3 \rangle_t = \mathrm{Cov}_t[x^2, x] + \langle x^2 \rangle_t \cdot \langle x \rangle_t.
\]
Multiplying by $t^2$:
\[
  t^2 \cdot \langle x^3 \rangle_t
  = t^2 \cdot \mathrm{Cov}_t[x^2, x]
    + (t \cdot \langle x^2 \rangle_t) \cdot (t \cdot \langle x \rangle_t).
\]
Each factor on the right has a known limit from Tide 9's bag of
asymptotics: $-2\alpha/\lambda^3$ for the covariance, $1/\lambda$ for
$t \cdot \langle x^2 \rangle$, and $-\alpha/(2\lambda^2)$ for
$t \cdot \langle x \rangle$. By \texttt{Filter.Tendsto.add} +
\texttt{Filter.Tendsto.mul}:
\[
  t^2 \cdot \langle x^3 \rangle_t
  \;\longrightarrow\;
  -\frac{2\alpha}{\lambda^3}
  + \frac{1}{\lambda} \cdot \left(-\frac{\alpha}{2\lambda^2}\right)
  = -\frac{5\alpha}{2\lambda^3}.
\]
The terminal arithmetic is one \texttt{field\_simp + ring}.

The Lean structure follows the proof verbatim:
1. Pull in the three asymptotic hypotheses
(\texttt{hCov}, \texttt{hM2}, \texttt{hM1}).
2. Build the sum-of-products via \texttt{hCov.add (hM2.mul hM1)} —
this gives the limit value
$-2\alpha/\lambda^3 + (1/\lambda) \cdot (-\alpha/(2\lambda^2))$.
3. Rewrite the limit value to $-5\alpha/(2\lambda^3)$ via a
\texttt{have h\_lim\_eq} + \texttt{field\_simp ; ring}.
4. Bridge the function: the integrand on the LHS
$t^2 \cdot \langle x^3 \rangle$ equals the integrand on the RHS
$t^2 \cdot \mathrm{Cov}[x^2, x] + (t \cdot \langle x^2 \rangle)(t \cdot \langle x \rangle)$
after unfolding \texttt{gibbsCov} and a small \texttt{x\^{}2 * x = x\^{}3}
function-extensionality nudge.
5. \texttt{Tendsto.congr'} closes the goal.

\section{Roadblocks and resolutions}

There were no roadblocks. The proof followed the deliberation plan
verbatim, and the build was clean on first pass after one minor lint
warning (\texttt{show} vs \texttt{change}, which the linter handles).

The only technical detail worth recording: Tide 9's
\texttt{cov\_anharmonic\_asymptotic} is stated with the derived limit
value $-2\alpha/\lambda^3$ (not in the
$-2\alpha/\lambda^3 + (1/\lambda)\cdot(-\alpha/(2\lambda^2))$ form),
so the limit-value rewrite step had to combine the two sub-asymptotics'
limits into the headline limit value via \texttt{ring} arithmetic. The
\texttt{nhds (a + b)}-vs-\texttt{nhds c} pattern that recurred in Tides 9
and 10 reappeared here: build the sum, name the equality, rewrite. Three
minutes of plumbing.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.} The standard \texttt{Tendsto}-algebra
glue (\texttt{add}, \texttt{mul}, \texttt{congr'},
\texttt{filter\_upwards}). Nothing exotic.

\paragraph{Laplace provided.} Three asymptotics from Tide 9 / the
original anharmonic stack: covariance, second moment, and
mean.\footnote{Concretely
\texttt{cov\_anharmonic\_asymptotic},
\texttt{secondMoment\_anharmonic\_asymptotic}, and
\texttt{mean\_anharmonic\_asymptotic}.}

\paragraph{What was new.} One theorem
(\texttt{thirdMoment\_anharmonic\_asymptotic}, $\sim$30 lines) and one
limit-value rewrite. No new helper lemmas were introduced. The
mathematical content was already in the Tide 9 retrospective; this
tide is pure packaging.

\section{Lessons}

\begin{itemize}
\item \textbf{Strict-improvement follow-ups left in retrospectives are
load-bearing.} Tide 9's retrospective named exactly this corollary,
with the closed-form value, with the proof outline, with the line-count
estimate. Today's tide consumed that named follow-up directly. The
\texttt{tide-pick} skill's emphasis on harvesting retrospectives'
Follow-up sections paid off: this tide cost almost nothing because
the deliberation had already happened.

\item \textbf{Skip the GPT consult for genuine packaging tides.} For a
follow-up that's already named in a prior retrospective with the
explicit closed form and proof outline, a fresh GPT-5.5~Pro consult is
overhead. The deliberation lives in the prior retrospective; the
current tide's job is execution. Worth landing in the tide skill: \emph{a
``packaging'' modifier where Step 2 is reduced to a single-candidate
log entry and direct Step 3 hand-off, when the candidate is a named
follow-up of a closed retrospective.}

\item \textbf{The same plumbing pattern recurs.} The three Tides 9, 10,
and now this G3 all share a structure: sum/product of two-three
existing \texttt{Tendsto} facts plus a limit-value \texttt{ring}
rewrite plus a \texttt{Tendsto.congr'} bridge. Worth abstracting once
the pattern fires a fourth time (the \texttt{gibbsCov\_symm} +
covariance-algebra refactor in I2 from the survey is exactly this).
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{Affine-observable third moment.} Generalise to
$t^2 \cdot \langle (a x + c)^3 \rangle_t \to \cdots$ via multilinearity
and the existing $\langle x \rangle, \langle x^2 \rangle$ asymptotics.
$\sim$30 extra lines. Pairs with G2 (affine-observable anharmonic
$\kappa_3$).
\item \emph{Higher-moment asymptotics.} The same template lifts to
$\langle x^4 \rangle, \langle x^5 \rangle$ and beyond — but the
sub-asymptotics for those moments are not yet exposed in laplace and
the leading orders shift. Sequenced: would build atop a
``\texttt{thirdMoment} via the rescaling identity'' refactor first.
\item \emph{Update the survey.} Mark G3 closed.
\end{itemize}

\section*{Acknowledgements}

This tide skipped the deliberation step entirely. The deliberation
was inherited verbatim from Tide 9's retrospective Follow-ups section,
which named the closed form, the proof outline, and the line-count
estimate. GPT-5.5~Pro contributed nothing to this tide directly —
its load-bearing contribution was at Tide 9's deliberation, where it
caught the Wick-coefficient error that exposed the correct
$5\alpha/2$ term.

\end{document}
