\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{newunicodechar}
\newunicodechar{ℝ}{\ensuremath{\mathbb{R}}}
\newunicodechar{ℓ}{\ensuremath{\ell}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{·}{\ensuremath{\cdot}}
\newunicodechar{²}{\ensuremath{^2}}
\newunicodechar{³}{\ensuremath{^3}}
\newunicodechar{φ}{\ensuremath{\varphi}}
\newunicodechar{ψ}{\ensuremath{\psi}}
\newunicodechar{χ}{\ensuremath{\chi}}
\newunicodechar{κ}{\ensuremath{\kappa}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{μ}{\ensuremath{\mu}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{α}{\ensuremath{\alpha}}
\newunicodechar{∫}{\ensuremath{\int}}
\newunicodechar{∂}{\ensuremath{\partial}}
\newunicodechar{≠}{\ensuremath{\neq}}
\newunicodechar{₀}{\ensuremath{_0}}

\newcommand{\userbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{blue!60}{DM:}} #1\end{quote}\medskip}
\newcommand{\claudebox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{orange!60!black}{Claude:}} #1\end{quote}\medskip}
\newcommand{\gptbox}[1]{\par\medskip\begin{quote}\small\textbf{\textcolor{green!50!black}{GPT-5.5 Pro:}} #1\end{quote}\medskip}

\title{Tide retrospective:\\\emph{cross-susceptibility derivative vanishes for the harmonic Gibbs}}
\author{Daniel Murfet (with Claude Opus 4.7)}
\date{6 May 2026}

\begin{document}

\sloppy
\maketitle

\begin{abstract}
A six-line Tide-loop excursion in the \texttt{laplace} seabed,
composing two prior tides into a downstream theorem neither delivers
alone:
$\partial_h \mathrm{Cov}_h(x, x)|_{h=0} = -t\cdot\kappa_3(x,x,x) = 0$
for the harmonic Gibbs $L = (\lambda/2) x^2$ with linear perturbation
$A = x$. Lives in \texttt{Laplace/OneD/HarmonicCrossSuscDeriv.lean}
(87 lines, including header docs); the proof body is six lines on top
of \texttt{harmonic\_id\_gibbsRegularity} (Tide 10),
\texttt{kappa3\_harmonic\_id\_id\_id\_eq\_zero} (Tide 5), and
threepoint's generic cross-susceptibility identity (Tide 6). Zero
\texttt{sorry}, zero \texttt{axiom}, zero \texttt{native\_decide}.
The interesting friction in this run was operational, not
mathematical: two other Tide agents (separable-potential and
bounded-prior-test-function) were operating concurrently in the same
checkout, and the \texttt{Laplace.lean} import line and the new file
itself were churned by branch-switching and staging interleavings,
including a commit that landed my file content under another agent's
draft commit message. Resolution: amend the bad message; carry the
\emph{per-Tide \texttt{git worktree}} recommendation forward (it had
already been raised by the 2D pure-quartic retrospective on 2~May).
\end{abstract}

\section{Setting}

The threepoint paper develops the susceptibility framework around
three load-bearing identities: the FDT
($\partial_h \langle\varphi\rangle_h = -t\,\mathrm{Cov}(\varphi, A)$),
the cross-susceptibility identity
($\partial_h \mathrm{Cov}_h(\varphi, B) = -t\,\kappa_3(\varphi, A, B)$),
and the four-point flow equation. The threepoint repo (Tide 6 + earlier)
proves these in the generic-Gibbs setting, parameterised by a
\texttt{GibbsRegularity} structure and per-observable \texttt{GibbsObservable}
hypotheses.

For the framework to deliver concrete theorems on a specific
potential, a project needs (i) the \texttt{GibbsRegularity} instance
for that potential, and (ii) \texttt{GibbsObservable} instances for the
observables of interest. Tide~10 supplied (i) for the harmonic +
linear case (\texttt{Threepoint.harmonic\_id\_gibbsRegularity}), and
the Tide~10 retrospective named the natural one-tide capstone:

\gptbox{Cross-susceptibility derivative theorems for the harmonic case.
With the regularity instance now in hand, the FDT and cross-susceptibility
derivatives specialise concretely. The cross-susceptibility derivative is
particularly satisfying: by Tide~5,
$\kappa_3(\varphi, x, B) = 0$ for the harmonic case, so
$\partial_h \mathrm{Cov}_h(\varphi, B)|_{h=0} = -t \cdot \kappa_3(\varphi, x, B) = 0$
for any $\varphi, B$ for which the relevant \texttt{GibbsObservable}
hypotheses hold. A natural one-tide step would package this as a stand-alone
theorem in laplace.}

The candidate (G1 in the May~6 survey) is that capstone. The Tide-pick
auto-proceed timer fired on G1; this Tide implements it.

The Step~2 deliberation\footnote{See \texttt{projects/patterning/tide-log/2026-05-06-tide-cross-susc-deriv-harmonic.md}.}
caught two issues with the survey's wording and an alternative route
worth recording.

\section{The thing formalised}

\begin{quote}\itshape
For the harmonic Gibbs measure
$\mu_{t,h}(x) \propto \exp(-t((\lambda/2)x^2 + h\,x))$ on $\mathbb{R}$
against Lebesgue, with $\lambda, t > 0$, and given
\texttt{GibbsObservable} hypotheses for the monomials $x$, $x^2$, and
$x^3$ at the harmonic + linear setup,
\[
  \frac{\partial}{\partial h} \mathrm{Cov}_h(x, x)\bigg|_{h=0} \;=\; 0.
\]
\end{quote}

\noindent The Lean signature, in namespace \texttt{Laplace.OneD}:
\begin{verbatim}
theorem cov_h_id_id_deriv_harmonic_eq_zero
    {lam t : ℝ} (hlam : 0 < lam) (ht : 0 < t)
    (hx  : Threepoint.GibbsObservable (volume : Measure ℝ)
            ((λ/2)·²) (id) t (id))
    (hx2 : Threepoint.GibbsObservable (volume : Measure ℝ)
            ((λ/2)·²) (id) t (id*id))
    (hx3 : Threepoint.GibbsObservable (volume : Measure ℝ)
            ((λ/2)·²) (id) t (id*id*id)) :
    HasDerivAt
        (fun h => Threepoint.gibbsCov (volume) ((λ/2)·²) (id) t h
                                       (id) (id))
        0 0
\end{verbatim}

\noindent (using \texttt{(λ/2)·²} as shorthand for \texttt{fun x : ℝ
=> lam / 2 * x ^ 2} and \texttt{id} for \texttt{fun x : ℝ => x}; the
file source spells these out).

The hypothesis pattern is that the three external \texttt{GibbsObservable}s
encode the differentiation-under-the-integral content for the three
monomials. Closing them is candidate G4 of the May~6 survey, deferred
here on grounds of scope.

The conclusion is in fact weaker than what holds. By square completion,
the perturbed Gibbs measure factors as
$\exp(t h^2/(2\lambda)) \cdot (\text{Gaussian centred at }-h/\lambda
\text{ with variance } 1/(\lambda t))$, so $\mathrm{Cov}_h(x, x) = 1/(\lambda
t)$ is \emph{globally} constant in $h$. The $\kappa_3$-route delivers
only the basepoint version of this fact; the global-constancy form is
recorded as a future tide.

\section{Proof strategy}

The proof is a structural composition of three prior results, with
\texttt{ring}-level arithmetic at the joins. Five lines of Lean tactics:

\begin{enumerate}[leftmargin=*]
\item Discharge \texttt{GibbsRegularity}\,via Tide 10's instance.
\item Apply threepoint's \texttt{gibbsCov\_deriv\_eq\_neg\_t\_kappa3}
with $(\varphi, A, B) = (\mathrm{id}, \mathrm{id}, \mathrm{id})$. The six
\texttt{GibbsObservable} slots ($\varphi$, $B$, $\varphi B$, $\varphi A$, $BA$,
$\varphi BA$) collapse to three distinct hypotheses (\texttt{hx} fills
$\varphi$ and $B$; \texttt{hx2} fills the three pairwise products; \texttt{hx3}
fills the triple). The conclusion is\linebreak
\texttt{HasDerivAt} of the covariance, with target derivative
$-t \cdot \kappa_3(\mathrm{id}, \mathrm{id}, \mathrm{id})$.
\item Rewrite the $\kappa_3$ term using Tide 5
(\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero}) to get
$-t \cdot 0$.
\item \texttt{mul\_zero} reduces to $0$.
\item \texttt{exact} the resulting hypothesis.
\end{enumerate}

The interesting structural feature is what does \emph{not} appear: no
calculus done by hand, no integration, no differentiation under the
integral, no parity argument carried out in this file. Tide 5 is
where the parity argument lives; Tide 10 is where the analytic
content of the partition function lives; this tide is just the gate
that admits both.

\section{Roadblocks and resolutions}

\paragraph{The retrospective's claim about Tide 5 was loose.} The
Tide 10 retrospective (which this Tide is the capstone of) says
``by Tide 5, $\kappa_3(\varphi, x, B) = 0$ for the harmonic case'',
suggesting Tide 5 covers general $(\varphi, x, B)$. It does not. Tide 5
covers only $(\mathrm{id}, \mathrm{id}, \mathrm{id})$. GPT-5.5 Pro
caught this immediately at the deliberation step and supplied a
concrete Gaussian counterexample for the general claim:

\gptbox{For a centred Gaussian with variance $\sigma^2$,
$\kappa_3(x^2, x, x) = \mathbb{E}[x^4] - \mathbb{E}[x^2]\mathbb{E}[x^2]
= 3\sigma^4 - \sigma^4 = 2\sigma^4 \neq 0$. So the honest G1 really is
the restricted $(\varphi, B) = (\mathrm{id}, \mathrm{id})$ version.}

The deliberation log records this counterexample so the next tide
that wants the general $(\varphi, x, B)$ statement starts from the
right premise.

\paragraph{No proof-side roadblocks beyond a whitespace warning.} The
six-line proof landed first try modulo a single linter warning (an
extra space before a colon in a hypothesis declaration). Lean's
elaborator handled the integrand unification at the
\texttt{GibbsObservable} slots automatically: the
$\beta$-equivalence between $(\mathrm{fun}\,w \mapsto (\mathrm{fun}\,x
\mapsto x)\,w \cdot (\mathrm{fun}\,x \mapsto x)\,w)$ and $(\mathrm{fun}\,x
\mapsto x \cdot x)$ went through without a \texttt{simpa} or
\texttt{show}; GPT had warned about possible
parenthesisation friction, which did not materialise.

\paragraph{Concurrent-Tide-agent collisions.} The dominant friction
in this run was operational, not mathematical. Two other Tide branches
were running in the same checkout
during the same window. Concrete symptoms observed:

\begin{itemize}
\item \emph{Import-line clobber.} My one-line addition to
\texttt{Laplace.lean} (the import for the new file) was overwritten
within the staging window: by the time the build started, another
agent had replaced my line with their own \texttt{QuarticBoundedPriorTestFn}
import. Re-add fixed it.
\item \emph{File disappearance on branch-switch.} The new
file\footnote{\texttt{HarmonicCrossSuscDeriv.lean}.} was untracked at
the moment when the working tree was switched between branches by
another agent; after the switch, the file was gone from disk.
Switching back to my branch restored it (because the file was
actually re-tracked on it under a stray commit).
\item \emph{Stray commit with wrong message.} A commit landed on
this Tide's branch\footnote{\texttt{tide/cross-susc-deriv-harmonic}.}
containing exactly my work (the new file plus a one-line
\texttt{Laplace.lean} addition, 88 lines total) but with the
bounded-prior agent's commit message. Resolution:
\texttt{git commit --amend} to fix the message; the content was
already correct. Reflog has the original.
\end{itemize}

The 2D pure-quartic retrospective (Tide step 5 in the laplace chain,
2~May) had already raised exactly this category of failure and named
the structural fix:

\gptbox{Concurrent Tide agents need filesystem isolation. The
\texttt{Laplace.lean} import-list is a contention point because every
new top-level file edits it. Two agents in the same checkout will
race; it is not a question of careful sequencing. The structural fix
is to run each Tide agent in its own \texttt{git worktree} attached
to its own branch.}

Today's run is one more data point. The fix is overdue.

\section{What was Mathlib, what was new}

\paragraph{Mathlib provided.} Nothing except \texttt{mul\_zero}. The
proof body is purely \texttt{ring}-level / \texttt{rw}-level
manipulation of an existing \texttt{HasDerivAt} hypothesis.

\paragraph{Threepoint and laplace provided.} All the infrastructure.
The three load-bearing prior tides:
\texttt{Threepoint.harmonic\_id\_gibbsRegularity} (Tide 10),
\texttt{Threepoint.kappa3\_harmonic\_id\_id\_id\_eq\_zero} (Tide 5),
\texttt{Threepoint.gibbsCov\_deriv\_eq\_neg\_t\_kappa3}
(Tide 6, the generic cross-susceptibility identity).

\paragraph{What was new.} The \emph{composition.} A theorem that
combines three pre-existing pieces and unconditionally produces a
named harmonic-case fact, with the three remaining
\texttt{GibbsObservable} hypotheses left as parameters. Not novel
mathematical content; novel \emph{packaging} that turns a
parameterised lemma into a theorem about a specific Gibbs measure.
This is exactly the bridge step the Tide loop is good at producing.

\section{Lessons}

\begin{itemize}
\item \emph{Concurrent Tide agents in the same checkout
remain a structural failure mode.} The 2~May retrospective named the
fix (\texttt{git worktree} per Tide branch); today's run is an
overdue reminder. Worth lifting into the Tide skill as a preflight
step that automatically creates a worktree, not just as advice.

\item \emph{Retrospectives leak loose claims.} The Tide 10
retrospective overstated Tide 5's content (``\,$\kappa_3(\varphi, x, B)=0$\,''
for general $(\varphi, B)$). A future tide that read the retrospective
and not the source might have wasted time on a false target. Discipline:
when a retrospective references a prior tide's theorem, copy the
\emph{Lean signature verbatim}, not a paraphrase. The deliberation
step caught this here, but only because GPT supplied a
counterexample; without the multi-model loop the loose claim might
have propagated.

\item \emph{Composition tides are honest tide steps.} A six-line
proof is a perfectly valid Tide outcome, provided the deliberation
step does the work of identifying the precise composition shape and
the right premises. The package-of-three-premises design here makes
the theorem's place in the ecosystem unambiguous: anyone who later
discharges the three \texttt{GibbsObservable} hypotheses
(candidate G4) gets the unconditional version in one extra line.

\item \emph{The numerical sanity check still earned its keep.} For a
``structural'' identity ($\partial_h \mathrm{Cov} = 0$) it would have
been tempting to skip the check. Doing it surfaced the stronger fact
that $\mathrm{Cov}_h$ is \emph{globally} constant ($1/(\lambda t)$),
not just locally constant at the basepoint. That observation
became the runner-up alternative GPT named, recorded as a future
tide candidate. A skipped check would have left it unnoticed.
\end{itemize}

\section{Follow-ups}

\begin{itemize}
\item \emph{G4: discharge the three \texttt{GibbsObservable} hypotheses.}
For each $k \in \{1, 2, 3\}$, prove
$\mathrm{Threepoint.GibbsObservable}\,(\mathrm{volume})\,((\lambda/2)\cdot^2)\,(\mathrm{id})\,t\,(\mathrm{fun}\,x \mapsto x^k)$.
Closed-form route via square completion + Gaussian moments
(per GPT's recommended derivation). Estimated $\sim$300--400 lines for
the three cases together; makes today's theorem unconditional.

\item \emph{Global-constancy strengthening (the runner-up).} Prove
$\mathrm{Threepoint.gibbsCov}\,(\mathrm{volume})\,((\lambda/2)\cdot^2)\,(\mathrm{id})\,t\,h\,(\mathrm{id})\,(\mathrm{id})\, =\, 1/(\lambda t)$
for all $h$, not just locally constant at $h = 0$. Direct via
shifted-Gaussian translation invariance plus the closed-form partition
already in Tide 10. Avoids \texttt{GibbsObservable} hypotheses
entirely. GPT's runner-up suggestion at deliberation; not chosen for
this Tide because the package shape was different.

\item \emph{Structural fix for concurrent-Tide-agents.} A wrapper
that runs \texttt{git worktree add} per Tide branch and points the
agent at the worktree. Doing this once removes the entire class of
collisions documented above. Not a Lean tide; a process tide for the
Tide skill itself.

\item \emph{Update the May 6 candidates survey.} Mark G1 as ✓ Closed.
The skill's Step 4 close-out section requires this.

\item \emph{\texttt{\textbackslash leanref}-tag the threepoint paper.}
The harmonic case now has three formalised theorems
(\texttt{kappa3\_harmonic\_id\_id\_id\_eq\_zero}, the
\texttt{GibbsRegularity} instance, and today's
\texttt{cov\_h\_id\_id\_deriv\_harmonic\_eq\_zero}), enough for a
one-pass tagging consistent with the laplace primer setup. This is
candidate H4 in the May 6 survey.
\end{itemize}

\section*{Acknowledgements}

GPT-5.5 Pro contributed the deliberation-stage advisory at Step 2,
including the explicit Gaussian counterexample to the general
$\kappa_3(\varphi, x, B) = 0$ claim and the runner-up
shifted-Gaussian alternative. The full consult is preserved at
\texttt{projects/patterning/tide-log/gpt55\_tide\_cross\_susc\_deriv\_harmonic\_v1.md}.

\end{document}
